---
title: "STAT432 Final Project: Predicting Online Shoppers' Purchasing Intention"
output: pdf_document
date: "2026-04-30"
---

\begin{center}
\textbf{Andrew Di} (amdi2) \quad \textbf{Praveen Vijayakumar} (pv16) \quad \textbf{Gautham Velupally} (gkv4)
\end{center}

### Abstract

In this report we explore the relation between the multiple factors about an online shopper and their purchasing intention — did they buy a product or not. We take a look at the UCI Online Shoppers Purchasing Intention dataset found at https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset. The dataset has $12{,}330$ observations and $18$ variables. We want to predict whether or not a customer bought an item (the `Revenue` variable) in a given session using the other variables as inputs. These other variables include things like the location of the shopper, the time of the year at which they are shopping, and their computer's profile, as well as information about the products they are viewing and the pages they are viewing.

In the studies we conducted, we handled the imbalance of the dataset, as well as fit multiple different types of models to determine the relationship between the 17 predictor variables in our dataset and whether or not a shopper purchased the product. For example, we fit K-Means Clustering and hierarchical clustering to determine the relation between the variables through unsupervised methods. We also fit multiple methods following supervised approaches, such as Linear Discriminant Analysis, Random Forests, and Boosted Trees.

Our conclusion is that `PageValues`, the average value of a page that a user visited before completing a transaction, seems to be the most important variable for predicting whether or not a shopper purchased a product, and a lot of the models perform similarly, but the Boosted Tree seems to have the best performance in terms of AUC.

*Note: AI tools were used to assist with portions of the code in this report.*

### Member Contributions

- **Andrew Di**: Data preprocessing, data visualization, and logistic regression.
- **Praveen Vijayakumar**: Tree-based methods (Random Forest, Gradient Boosting), log-transformation of features, and variable importance analysis.
- **Gautham Velupally**: PCA, K-means and hierarchical clustering, LDA, KNN

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Introduction and Contextualization

"Online Shoppers Purchasing Intention Dataset" provided by UCI Machine Learning Repository [here](https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset).

The dataset contains $12{,}330$ observations of online shopping sessions, with $17$ input variables and one categorical output variable, `Revenue`, indicating whether the session ended in a purchase. The classes are imbalanced, with only about $15\%$ of sessions resulted in a purchase. There is no missing data.

We divide the $17$ predictors into four conceptual groups:

- **Page Interaction Features**: the number of pages of each type visited in the session and the total time spent on them: `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, and `ProductRelated_Duration`. These are derived from URL information and updated in real time as the user moves between pages.

- **Google Analytics Metrics** — engagement-quality metrics from Google Analytics: `BounceRates` (the percentage of visitors who enter the site from a page and leave without triggering any further requests), `ExitRates` (the percentage of pageviews on a page that were the last in the session), and `PageValues` (the average value of a page that a user visited before completing a transaction).

- **Session Context Features** — time- and seasonality-related variables: `SpecialDay`, `Month`, and `Weekend`. The `SpecialDay` value reflects how close the visit is to a holiday such as Valentine's Day or Mother's Day; for Valentine's Day, for instance, it takes a nonzero value between February 2 and February 12 and peaks at $1$ on February 8.

- **User Technical Profile** — session-level user attributes: `OperatingSystems`, `Browser`, `Region`, `TrafficType`, and `VisitorType` (whether the user is a new or returning visitor).

Our goal is to build classification models that can effectively determine whether or not an online shopper will make a purchase based on the feature variables above.

## Data Preprocessing and Visualization

We'll start by loading in our data.

```{r load-data}
library(tidyverse)

shoppers <- read_csv("online_shoppers_intention.csv")
glimpse(shoppers)
```

Before fitting any models, we want to inspect the distribution of each input variable individually. This helps us visually identify features that may be skewed or that may follow a distribution that we can take advantage of. We separate variables into numeric and categorical groups, plotting histograms for numeric features and bar plots for categorical features.

```{r univariate-numeric, fig.width=9, fig.height=5}
numeric_features <- c(
  "Administrative", "Administrative_Duration",
  "Informational", "Informational_Duration",
  "ProductRelated", "ProductRelated_Duration",
  "BounceRates", "ExitRates", "PageValues", "SpecialDay"
)

shoppers %>%
  select(all_of(numeric_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  labs(title = "Distributions of Numeric Features", x = NULL, y = "Count") +
  theme_minimal()
```

After plotting, we notice that each of our numeric features is mostly centered near $0$ and have skew to the right. We may want to perform some log transformation of our features to handle the skew. Notably, `SpecialDay` contains discrete values which may require special handling.

```{r univariate-categorical, fig.width=9, fig.height=5}
categorical_features <- c(
  "Month", "OperatingSystems", "Browser",
  "Region", "TrafficType", "VisitorType", "Weekend"
)

shoppers %>%
  select(all_of(categorical_features)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  labs(title = "Distributions of Categorical Features", x = NULL, y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Several of the categorical distributions are heavily imbalanced. `Browser`, `OperatingSystems`, and `TrafficType` are dominated by one or two levels with a long tail of rare ones, which suggests we may want to lump rare levels together (e.g. via `forcats::fct_lump`) before modeling so that the rare levels don't introduce instability. `VisitorType` is mostly returning visitors with a smaller share of new visitors, and `Weekend` is mostly `FALSE`. `Month` shows a seasonal pattern with clear peaks (notably in May and November). Whether features like `OperatingSystems` or `Browser` carry real predictive signal is an open question — we'll get a better sense of that in the bivariate plots below.

### Bivariate Distributions vs. Revenue

Univariate plots tell us how each feature is distributed in isolation, but for a classification problem the more useful question is: *does this feature differ between sessions that ended in a purchase and those that did not?* If the conditional distributions of a feature given `Revenue = TRUE` and `Revenue = FALSE` look similar, that feature carries little signal; if they look different, it likely matters. We use boxplots for the numeric features (comparing distributions by `Revenue`) and stacked proportion bars for the categorical features (showing the proportion of purchases within each level relative to the $\sim 15\%$ overall purchase rate).

```{r bivariate-numeric, fig.width=9, fig.height=5}
shoppers %>%
  select(all_of(numeric_features), Revenue) %>%
  pivot_longer(-Revenue, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = Revenue, y = value, fill = Revenue)) +
  geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  labs(title = "Numeric Features by Revenue", x = "Revenue", y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r bivariate-categorical, fig.width=9, fig.height=5}
shoppers %>%
  select(all_of(categorical_features), Revenue) %>%
  mutate(across(-Revenue, as.character)) %>%
  pivot_longer(-Revenue, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = Revenue)) +
  geom_bar(position = "fill") +
  geom_hline(yintercept = mean(shoppers$Revenue), linetype = "dashed", color = "black") +
  facet_wrap(~ feature, scales = "free_x", ncol = 3) +
  labs(title = "Purchase Proportion by Categorical Feature",
       subtitle = "Dashed line marks the overall purchase rate (~15%)",
       x = NULL, y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Several of our numeric features are likely to be strongly related by construction — for example, the count of pages of a given type and the total time spent on those pages should rise together, and `BounceRates` and `ExitRates` measure closely related notions of leaving the site. Highly collinear predictors are a problem for linear models like logistic regression (unstable coefficients, inflated standard errors) and motivate the use of regularization methods like Lasso or Ridge. A scatterplot matrix (see Appendix, Figure A1) confirms this: each count-and-duration pair lies almost exactly on a line, and `BounceRates` and `ExitRates` show a strong positive linear relationship. Beyond these tight pairs, the relationships are weaker. Most features exhibit heavy right-skew (points clustered near the origin), reinforcing the case for log-transforming before fitting models.

## Data Preparation for Modeling

Before fitting any models, we need to encode categorical variables, split the data into training and test sets, and standardize numeric features. We convert `Revenue` to a factor, encode `Month` and `VisitorType` as dummy variables, and treat the integer-coded nominal features (`OperatingSystems`, `Browser`, `Region`, `TrafficType`) as factors to be dummy-encoded as well.

```{r data-prep, message=FALSE}
library(caret)
library(MASS)
library(class)
library(e1071)
library(cluster)

set.seed(432)

shoppers_model <- shoppers %>%
  mutate(
    Revenue = factor(Revenue, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    Month = factor(Month),
    VisitorType = factor(VisitorType),
    Weekend = factor(Weekend),
    OperatingSystems = factor(OperatingSystems),
    Browser = factor(Browser),
    Region = factor(Region),
    TrafficType = factor(TrafficType)
  )

dummy_formula <- dummyVars(Revenue ~ ., data = shoppers_model, fullRank = TRUE)
shoppers_encoded <- predict(dummy_formula, newdata = shoppers_model) %>% as.data.frame()
shoppers_encoded$Revenue <- shoppers_model$Revenue

train_idx <- createDataPartition(shoppers_encoded$Revenue, p = 0.7, list = FALSE)
train_data <- shoppers_encoded[train_idx, ]
test_data <- shoppers_encoded[-train_idx, ]

train_x <- train_data %>% dplyr::select(-Revenue)
train_y <- train_data$Revenue
test_x <- test_data %>% dplyr::select(-Revenue)
test_y <- test_data$Revenue

nzv_cols <- nearZeroVar(train_x, names = TRUE)
if (length(nzv_cols) > 0) {
  train_x <- train_x[, !names(train_x) %in% nzv_cols]
  test_x <- test_x[, !names(test_x) %in% nzv_cols]
}

preproc <- preProcess(train_x, method = c("center", "scale"))
train_x_scaled <- predict(preproc, train_x)
test_x_scaled <- predict(preproc, test_x)
```

## Unsupervised Learning

### Principal Component Analysis (PCA)

We apply PCA to the 10 original numeric features (before dummy encoding) to understand the variance structure and to provide reduced-dimension inputs for later models.

```{r pca, fig.width=9, fig.height=3.5}
train_numeric <- train_data %>% dplyr::select(all_of(numeric_features))
test_numeric <- test_data %>% dplyr::select(all_of(numeric_features))

pca_preproc <- preProcess(train_numeric, method = c("center", "scale"))
train_numeric_scaled <- predict(pca_preproc, train_numeric)
test_numeric_scaled <- predict(pca_preproc, test_numeric)

pca_fit <- prcomp(train_numeric_scaled, center = FALSE, scale. = FALSE)

var_explained <- summary(pca_fit)$importance
par(mfrow = c(1, 2))
plot(var_explained[2, ], type = "b", pch = 19, col = "steelblue",
     xlab = "Principal Component", ylab = "Proportion of Variance",
     main = "Scree Plot")
plot(var_explained[3, ], type = "b", pch = 19, col = "darkred",
     xlab = "Principal Component", ylab = "Cumulative Proportion",
     main = "Cumulative Variance Explained")
abline(h = 0.9, lty = 2, col = "gray40")
par(mfrow = c(1, 1))
```

```{r pca-biplot, fig.width=7, fig.height=4}
train_pca_scores <- as.data.frame(pca_fit$x[, 1:2])
train_pca_scores$Revenue <- train_y

ggplot(train_pca_scores, aes(x = PC1, y = PC2, color = Revenue)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "PCA: First Two Principal Components Colored by Revenue",
       x = paste0("PC1 (", round(var_explained[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2, 2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "gray60", "Yes" = "tomato"))
```

The PCA biplot shows that the two classes overlap substantially in the first two principal components, indicating that separating purchasers from non-purchasers requires more than two linear directions. However, `Revenue = Yes` sessions tend to have a wider spread, particularly along PC1, suggesting that high-engagement sessions (which load onto the page count and duration variables) are more likely to result in purchases.

### K-Means Clustering

We cluster the scaled numeric features to discover natural session groupings. We use the silhouette method to choose the number of clusters.

```{r kmeans-silhouette, fig.width=7, fig.height=5}
set.seed(432)
sample_idx <- sample(nrow(train_numeric_scaled), 3000)
train_sample <- train_numeric_scaled[sample_idx, ]

sil_widths <- sapply(2:8, function(k) {
  km <- kmeans(train_sample, centers = k, nstart = 25, iter.max = 50)
  ss <- silhouette(km$cluster, dist(train_sample))
  mean(ss[, 3])
})

plot(2:8, sil_widths, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (k)", ylab = "Average Silhouette Width",
     main = "Silhouette Analysis for K-Means")
best_k <- which.max(sil_widths) + 1
```

```{r kmeans-fit, fig.width=9, fig.height=5}
km_fit <- kmeans(train_numeric_scaled, centers = best_k, nstart = 25, iter.max = 100)

train_pca_all <- as.data.frame(pca_fit$x[, 1:2])
train_pca_all$Cluster <- factor(km_fit$cluster)
train_pca_all$Revenue <- train_y

p1 <- ggplot(train_pca_all, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "K-Means Clusters in PCA Space") +
  theme_minimal()

cluster_revenue <- train_pca_all %>%
  group_by(Cluster) %>%
  summarise(
    n = n(),
    purchase_rate = mean(Revenue == "Yes"),
    .groups = "drop"
  )

p2 <- ggplot(cluster_revenue, aes(x = Cluster, y = purchase_rate, fill = Cluster)) +
  geom_col() +
  geom_hline(yintercept = mean(train_y == "Yes"), linetype = "dashed") +
  labs(title = "Purchase Rate by Cluster",
       subtitle = "Dashed line = overall training rate",
       y = "Purchase Rate") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

```{r kmeans-profiles}
cluster_profiles <- train_data %>%
  dplyr::select(all_of(numeric_features)) %>%
  mutate(Cluster = factor(km_fit$cluster)) %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean), .groups = "drop")

cluster_profiles_long <- cluster_profiles %>%
  pivot_longer(-Cluster, names_to = "Feature", values_to = "Mean") %>%
  pivot_wider(names_from = Cluster, values_from = Mean, names_prefix = "Cluster ")
```

Cluster centroid profiles and a hierarchical clustering dendrogram corroborating these groupings are provided in the Appendix (Table A1 and Figure A2).

## Supervised Learning

We now turn to classification. For all supervised models, we evaluate using 5-fold stratified cross-validation on the training set. Because the classes are imbalanced ($\sim 85\%$ non-purchase), we report not just accuracy but also sensitivity (recall for purchases), specificity, and AUC. We collect results into a summary table at the end.

```{r cv-setup}
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

results <- list()
```

### Discriminant Analysis: Choosing Between LDA, QDA, and Naive Bayes

Among generative classifiers, we consider three options: Linear Discriminant Analysis (LDA), Quadratic Discriminant Analysis (QDA), and Naive Bayes. We select **LDA** for the following reasons:

- **QDA** requires estimating a separate covariance matrix for each class. With our high-dimensional dummy-encoded feature space, the within-class covariance matrices become singular or near-singular, forcing us to restrict QDA to a small PCA subspace and lose information from the categorical predictors.
- **Naive Bayes** assumes conditional independence among features given the class — a strong assumption clearly violated here (e.g., `BounceRates` and `ExitRates` are highly correlated, as are page counts and durations). While Naive Bayes can still produce reasonable rankings, its probability estimates are poorly calibrated under dependence, and it typically underperforms LDA when the shared-covariance assumption is approximately satisfied.
- **LDA** assumes multivariate normality with a shared covariance matrix across classes. While normality is imperfect for our skewed features, the shared-covariance assumption is more defensible than independence, and LDA can operate on the full encoded feature set without dimensionality issues. It also provides a natural projection onto the discriminant axis, aiding interpretation.

### Linear Discriminant Analysis (LDA)

LDA projects the data onto the single linear discriminant axis that maximally separates the two classes. Given the collinearity we observed earlier, we should expect LDA to perform comparably to logistic regression.

```{r lda-fit}
lda_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "lda",
  trControl = ctrl,
  metric = "ROC"
)

lda_pred <- predict(lda_model, test_x_scaled)
lda_prob <- predict(lda_model, test_x_scaled, type = "prob")

lda_cm <- confusionMatrix(lda_pred, test_y, positive = "Yes")

results[["LDA"]] <- list(
  model = lda_model,
  cm = lda_cm,
  probs = lda_prob
)
```

### K-Nearest Neighbors (KNN)

KNN classifies each observation by a majority vote of its $k$ nearest neighbors. It is a non-parametric, lazy-learning method — no model is fitted at training time; all computation happens at prediction time. This makes KNN impractical for real-time session scoring (as noted in the original paper), but it provides a useful non-parametric reference. We tune $k$ via cross-validation. Given the curse of dimensionality, we also try KNN on PCA-reduced features.

```{r knn-fit}
knn_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pred <- predict(knn_model, test_x_scaled)
knn_prob <- predict(knn_model, test_x_scaled, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_y, positive = "Yes")

results[["KNN"]] <- list(
  model = knn_model,
  cm = knn_cm,
  probs = knn_prob
)
```

```{r knn-pca, fig.width=7, fig.height=5}
train_pca_full <- as.data.frame(pca_fit$x[, 1:7])
test_pca_full <- as.data.frame(predict(pca_fit, test_numeric_scaled)[, 1:7])

knn_pca_model <- train(
  x = train_pca_full,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pca_pred <- predict(knn_pca_model, test_pca_full)
knn_pca_prob <- predict(knn_pca_model, test_pca_full, type = "prob")

knn_pca_cm <- confusionMatrix(knn_pca_pred, test_y, positive = "Yes")

results[["KNN (PCA)"]] <- list(
  model = knn_pca_model,
  cm = knn_pca_cm,
  probs = knn_pca_prob
)
```

## Tree Methods: Bagging, Boosting, and Random Forest

We now consider tree-based classifiers. First, we apply a log-transform to all numeric features. As the scatterplot matrix showed, nearly all numeric variables exhibit heavy right-skew with most observations clustered near zero. Skewed inputs can destabilize split selection in trees and inflate the influence of extreme values. Applying `log1p` (i.e., $\log(1+x)$) compresses the right tail while handling zeros gracefully, since $\log(0)$ is undefined but $\log(1) = 0$.

```{r log-transform}
train_x_log <- train_x %>%
  mutate(across(all_of(intersect(numeric_features, names(train_x))), log1p))
test_x_log <- test_x %>%
  mutate(across(all_of(intersect(numeric_features, names(train_x))), log1p))
```

### Random Forest

We fit a random forest with 1,000 trees and tune `mtry` (the number of features considered at each split) using out-of-bag (OOB) error. OOB error is the natural evaluation metric for bagged models: each tree is trained on a bootstrap sample, and the observations left out of that sample provide a built-in validation set without needing a separate cross-validation loop.

```{r rf-fit, results='hide'}
library(randomForest)

ctrl_oob <- trainControl(method = "oob", classProbs = TRUE)
rf <- train(
  x = train_x_log, y = train_y, method = "rf",
  trControl = ctrl_oob,
  tuneGrid = expand.grid(mtry = c(5, 10, 15, 20, ncol(train_x_log))),
  ntree = 1000, importance = TRUE
)
```

```{r rf-results}
knitr::kable(rf$results[, c("mtry", "Accuracy", "Kappa")], digits = 4,
             caption = "Random Forest: OOB Performance by mtry")
```

The optimal `mtry` is `r rf$bestTune$mtry`, achieving an OOB accuracy of `r round(rf$results$Accuracy[which.min(rf$results$Accuracy == max(rf$results$Accuracy))], 4)`. Note that the bagging model (`mtry` = `r ncol(train_x_log)`, using all predictors) performs worse than the random forest, confirming that decorrelating the trees by restricting the feature set at each split improves ensemble performance.

```{r rf-test, fig.height=4, fig.width=10}
rf_pred <- predict(rf, newdata = test_x_log)
rf_cm <- confusionMatrix(rf_pred, test_y, positive = "Yes")

varImpPlot(rf$finalModel, main = "Variable Importance for Online Shoppers")
```

`PageValues` is by far the most important variable by both mean decrease in accuracy and mean decrease in Gini, consistent with its role as a direct proxy for purchase intent. The test accuracy is `r round(rf_cm$overall["Accuracy"], 4)`.

### Gradient Boosting (GBM)

Unlike random forests, boosting builds trees sequentially, with each tree correcting the errors of the previous ensemble. We tune the number of trees, interaction depth, and learning rate via 5-fold cross-validation on accuracy. We use accuracy rather than OOB error because boosting does not employ bootstrap aggregation.

```{r gbm-fit, results='hide'}
gbm_grid <- expand.grid(
  n.trees = c(500, 1000),
  interaction.depth = c(3, 4),
  shrinkage = c(0.01, 0.1),
  n.minobsinnode = 10
)

ctrl_accuracy <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE, summaryFunction = defaultSummary
)

combined <- as.data.frame(train_x_log)
combined$Revenue <- train_y

gbm_fit <- train(
  Revenue ~ ., data = combined, method = "gbm",
  trControl = ctrl_accuracy, tuneGrid = gbm_grid,
  metric = "Accuracy", verbose = FALSE
)
```

```{r gbm-results}
best_gbm <- gbm_fit$bestTune
cat("Best GBM: n.trees =", best_gbm$n.trees,
    ", depth =", best_gbm$interaction.depth,
    ", shrinkage =", best_gbm$shrinkage, "\n")

gbm_imp <- summary(gbm_fit$finalModel, plotit = FALSE)
knitr::kable(head(gbm_imp, 10), digits = 2,
             caption = "GBM: Top 10 Variables by Relative Influence")
```

Again, `PageValues` dominates the relative influence rankings. The optimal model uses `r best_gbm$n.trees` trees with interaction depth `r best_gbm$interaction.depth` and a learning rate of `r best_gbm$shrinkage`.

```{r gbm-test}
gbm_pred <- predict(gbm_fit, newdata = test_x_log)
gbm_cm <- confusionMatrix(gbm_pred, test_y, positive = "Yes")
```

The GBM test accuracy is `r round(gbm_cm$overall["Accuracy"], 4)`, slightly higher than the random forest.

```{r tree-store-results}
results[["Random Forest"]] <- list(
  cm = rf_cm,
  probs = predict(rf, newdata = test_x_log, type = "prob")
)

results[["GBM"]] <- list(
  cm = gbm_cm,
  probs = predict(gbm_fit, newdata = test_x_log, type = "prob")
)
```

## Model Comparison

We now compare all models on key metrics computed from their test set predictions.

```{r model-comparison, fig.width=9, fig.height=5}
library(pROC)

comparison <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  F1 = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (name in names(results)) {
  cm <- results[[name]]$cm
  probs <- results[[name]]$probs

  roc_obj <- roc(test_y, probs$Yes, levels = c("No", "Yes"), quiet = TRUE)

  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1 <- ifelse(is.na(prec) | (sens + prec) == 0, 0, 2 * sens * prec / (sens + prec))

  comparison <- rbind(comparison, data.frame(
    Model = name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(sens, 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1 = round(f1, 4),
    AUC = round(auc(roc_obj), 4),
    stringsAsFactors = FALSE
  ))
}

rownames(comparison) <- NULL
knitr::kable(comparison, caption = "Model Comparison on Test Set")
```

```{r roc-curves, fig.width=8, fig.height=4}
roc_list <- list()
for (name in names(results)) {
  roc_list[[name]] <- roc(test_y, results[[name]]$probs$Yes,
                          levels = c("No", "Yes"), quiet = TRUE)
}

plot(roc_list[[1]], col = 1, main = "ROC Curves: Model Comparison")
for (i in 2:length(roc_list)) {
  lines(roc_list[[i]], col = i)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=", 
                       sapply(roc_list, function(r) round(auc(r), 3)), ")"),
       col = 1:length(roc_list), lwd = 2, cex = 0.8)
```

The comparison table and ROC curves reveal several patterns. The tree-based ensemble methods (GBM and Random Forest) achieve the highest accuracy and AUC, benefiting from their ability to capture nonlinear interactions and handle the mixed feature space. LDA performs well among the simpler methods due to the strong linear signal in `PageValues` and `ExitRates`. All models show relatively low sensitivity, a direct consequence of the $\sim 85\%$ class imbalance: with a default 0.5 threshold, classifiers are conservative and rarely predict a purchase unless the evidence is overwhelming. KNN suffers from the curse of dimensionality, though PCA-reduced features partially mitigate this. Across all methods, `PageValues` consistently emerges as the dominant predictor of purchase intent.

## Conclusion

Among the models evaluated, Gradient Boosted Trees achieved the best overall test performance, followed closely by Random Forest. Both ensemble methods substantially outperformed the simpler classifiers (LDA, KNN) in accuracy and AUC, owing to their capacity for modeling nonlinear feature interactions and their robustness to the mixed numeric-categorical feature space. We recommend GBM as the primary classifier for this task, with Random Forest as a strong alternative that offers more interpretable variable importance measures. Regardless of method, the class imbalance remains the central challenge: all models exhibit high specificity but limited sensitivity, indicating that threshold tuning or oversampling strategies would be necessary for deployment in a real-time purchase-intent flagging system.

### Future Work

From the tree models, the variable that is most important across all methods — both in mean decrease in accuracy and mean decrease in Gini — is `PageValues`. Conversely, the least important variables include several `TrafficType` levels, `Weekend`, `Month`, operating system indicators, and some `Region` levels. This suggests that temporal and geographic attributes do not contribute substantially to predicting revenue, nor does the operating system used by a potential customer. These findings are largely unsurprising, with the notable exception of `Month`: one might expect seasonality to play a significant role in purchasing behavior, given that different months bring different occasions and shopping surges. The relatively low importance of `Month` warrants further investigation — for instance, whether its effect is absorbed by `SpecialDay` or whether a more granular temporal encoding (e.g., week-of-year) would reveal stronger seasonal signal.

\newpage

## Appendix

### Figure A1: Scatterplot Matrix

```{r scatterplot-matrix, fig.width=10, fig.height=10}
shoppers_numeric <- subset(shoppers, select = numeric_features)
pairs(shoppers_numeric,
      main = "Scatterplot Matrix for Numeric Features",
      col = "navy",
      pch = 20)
```

### Table A1: Cluster Centroids

```{r appendix-cluster-profiles}
knitr::kable(cluster_profiles_long, digits = 2, caption = "Cluster Centroids (Original Scale)")
```

### Figure A2: Hierarchical Clustering Dendrogram

```{r appendix-hierarchical, fig.width=9, fig.height=5}
hc_dist <- dist(train_sample)
hc_fit <- hclust(hc_dist, method = "ward.D2")

plot(hc_fit, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method, n=3000 sample)",
     xlab = "", sub = "")
rect.hclust(hc_fit, k = best_k, border = "steelblue")
```