---
title: "STAT432 Final Project: Predicting Online Shoppers' Purchasing Intention"
output: pdf_document
date: "2026-04-30"
---

\begin{center}
\textbf{Andrew Di} (amdi2) \quad \textbf{Praveen Vijayakumar} (pv16) \quad \textbf{Gautham Velupally} (gkv4)
\end{center}

### Abstract

In this report we explore the relation between the multiple factors about an online shopper and their purchasing intention — did they buy a product or not. We take a look at the UCI Online Shoppers Purchasing Intention dataset found at https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset. The dataset has $12{,}330$ observations and $18$ variables. We want to predict whether or not a customer bought an item (the `Revenue` variable) in a given session using the other variables as inputs. These other variables include things like the location of the shopper, the time of the year at which they are shopping, and their computer's profile, as well as information about the products they are viewing and the pages they are viewing.

In the studies we conducted, we handled the imbalance of the dataset, as well as fit multiple different types of models to determine the relationship between the 17 predictor variables in our dataset and whether or not a shopper purchased the product. For example, we fit K-Means Clustering and hierarchical clustering to determine the relation between the variables through unsupervised methods. We also fit multiple methods following supervised approaches, such as Linear Discriminant Analysis, Random Forests, and Boosted Trees.

Our conclusion is that `PageValues`, the average value of a page that a user visited before completing a transaction, seems to be the most important variable for predicting whether or not a shopper purchased a product, and a lot of the models perform similarly, but the Boosted Tree seems to have the best performance in terms of AUC.

*Note: AI tools were used to assist with portions of the code in this report.*

### Member Contributions

- **Andrew Di**: Data preprocessing, data visualization, and logistic regression.
- **Praveen Vijayakumar**: Tree-based methods (Random Forest, Gradient Boosting), log-transformation of features, and variable importance analysis.
- **Gautham Velupally**: PCA, K-means and hierarchical clustering, LDA, KNN

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

**Team 13**

**Group Members:**

Andrew Di, amdi2
Praveen
Gautham

## Abstract {-}

[TODO: write a brief abstract of at most 200 words. Should cover (1) the prediction task and dataset, (2) the modeling approaches compared, (3) the headline result on the test set, and (4) the recommended approach.]

## Member Contributions {-}

Andrew Di: Data Preprocessing and Visualization, Logistic Regression
Praveen
Gautham

\newpage

# Introduction of the Task and Literature Review

"Online Shoppers Purchasing Intention Dataset" provided by UCI Machine Learning Repository [here](https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset).

The dataset contains $12{,}330$ observations of online shopping sessions, with $17$ input variables and one categorical output variable, `Revenue`, indicating whether the session ended in a purchase. The classes are imbalanced, with only about $15\%$ of sessions resulted in a purchase. There is no missing data.

We divide the $17$ predictors into four conceptual groups:

- **Page Interaction Features**: the number of pages of each type visited in the session and the total time spent on them: `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, and `ProductRelated_Duration`. These are derived from URL information and updated in real time as the user moves between pages.

- **Google Analytics Metrics** — engagement-quality metrics from Google Analytics: `BounceRates` (the percentage of visitors who enter the site from a page and leave without triggering any further requests), `ExitRates` (the percentage of pageviews on a page that were the last in the session), and `PageValues` (the average value of a page that a user visited before completing a transaction).

- **Session Context Features** — time- and seasonality-related variables: `SpecialDay`, `Month`, and `Weekend`. The `SpecialDay` value reflects how close the visit is to a holiday such as Valentine's Day or Mother's Day; for Valentine's Day, for instance, it takes a nonzero value between February 2 and February 12 and peaks at $1$ on February 8.

- **User Technical Profile** — session-level user attributes: `OperatingSystems`, `Browser`, `Region`, `TrafficType`, and `VisitorType` (whether the user is a new or returning visitor).

Our goal is to build classification models that can effectively determine whether or not an online shopper will make a purchase based on the feature variables above.

## Literature Review

[TODO: brief review of prior work, including the original paper introducing this dataset (Sakar et al., 2018, *Neural Computing and Applications*) and any related work on session-level e-commerce conversion prediction or imbalanced binary classification techniques relevant to the methods we use.]

# Details of the Dataset, Summary Statistics, and Data Visualization

## Data Preprocessing and Visualization

We'll start by loading in our data.

```{r load-data}
library(tidyverse)

# load data
shoppers <- read_csv("online_shoppers_intention.csv")
```

After loading in the data, we'll check for any missing data and print a summary of our dataset.

```{r clean}
numeric_features <- c(
  "Administrative", "Administrative_Duration",
  "Informational", "Informational_Duration",
  "ProductRelated", "ProductRelated_Duration",
  "BounceRates", "ExitRates", "PageValues", "SpecialDay"
)
categorical_features <- c(
  "Month", "OperatingSystems", "Browser",
  "Region", "TrafficType", "VisitorType", "Weekend"
)

summary(shoppers[, numeric_features])
```

We find that we don't have any missing data values. Initially, we notice that most of our numeric features are concentrated near lower values but seem to have a right skew. We'll plot our features individually to visualize and identify any skewness or distributions we can take advantage of. We separate variables into numeric and categorical groups, plotting histograms for numeric features and bar plots for categorical features.

```{r univariate-numeric, fig.width=9, fig.height=5}
par(mfrow = c(4, 3), mar = c(4, 4, 2, 1))
for (feat in numeric_features) {
  hist(shoppers[[feat]],
       breaks = 40,
       col = "steelblue",
       border = "white",
       main = feat,
       xlab = NULL)
}
par(mfrow = c(1, 1))
```

After plotting, we confirm that each of our numeric features is mostly concentrated at low values and skewed to the right. We may want to perform some log transformation of our features to handle the skew. Notably, `SpecialDay` contains discrete values which may require special handling.

```{r univariate-categorical, fig.width=9, fig.height=5}
par(mfrow = c(3, 3), mar = c(6, 4, 2, 1))
for (feat in categorical_features) {
  counts <- table(as.character(shoppers[[feat]]))
  barplot(counts,
          col = "steelblue",
          main = feat,
          las = 2,
          cex.names = 0.8)
}
par(mfrow = c(1, 1))
```

After plotting, we notice that several of our categorical features are heavily imbalanced. `Browser`, `OperatingSystems`, and `TrafficType` are mostly dominated by one or two levels. Similarly, `VisitorType` is mostly returning visitors with a smaller number of new visitors, and `Weekend` is mostly `FALSE`. `Month` shows a lot of seasonality with clear peaks in May and November, which could be explained by the end of the school year and graduation gifts in May, and the upcoming holiday season in November.

While plotting our features individually helps us understand their distribution, plotting the features against `Revenue` helps us identify if our feature distribution is dependent on whether or not a user made a purchase. If the conditional distributions of a feature given different `Revenue` value look different, then those features could carry more signal and are worth examining more. Below, we create box plots for our numeric features (compared against `Revenue`), and bar plots measuring per-level purchase rates for our categorical features. In our bar plots, the dashed line references the $\sim 15\%$ purchase rate in our overall dataset to compare against.

```{r bivariate-numeric, fig.width=9, fig.height=5}
par(mfrow = c(4, 3), mar = c(3, 4, 2, 1))
for (feat in numeric_features) {
  boxplot(shoppers[[feat]] ~ shoppers$Revenue,
          col = c("salmon", "steelblue"),
          main = feat,
          xlab = "Revenue",
          ylab = NULL,
          outcex = 0.4)
}
par(mfrow = c(1, 1))
```

When visually inspecting the conditional distribution of our numeric features against `Revenue`, we find that when `Revenue` is `TRUE`, our `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, `ProductRelated_Duration`, and `PageValues` tend to be higher, while `BounceRates` and `ExitRates` tend to be lower. Because of the distinct distributions when compared against different values of `Revenue`, these might be features that are more important in our models.

Additionally, each of our box plots contains some amount of outliers. We believe the heavy right skewness of our features makes sense. For example, most users probably don't spend that much time on a page, which is why `Administrative_Duration` and `Informational_Duration` are mostly concentrated at low values. However, a few users will certainly spend far more time on a page. We decided to keep the outliers for each of our numeric features because we believe they capture and represent the skewed behavior that we've described.

```{r bivariate-categorical, fig.width=9, fig.height=5}
overall_rate <- mean(shoppers$Revenue)
par(mfrow = c(3, 3), mar = c(6, 4, 2, 1))
for (feat in categorical_features) {
  rates <- tapply(shoppers$Revenue, as.character(shoppers[[feat]]), mean)
  barplot(rates,
          col = "steelblue",
          main = feat,
          ylab = "Purchase Rate",
          las = 2,
          cex.names = 0.8,
          ylim = c(0, max(c(rates, overall_rate)) * 1.1))
  abline(h = overall_rate, lty = 2, col = "red")
}
par(mfrow = c(1, 1))
```

Now, when examining our categorical features, we can identify certain levels in our features that correspond with a larger purchase rate than average. For example, the purchase rate on the weekend is higher than average, and new visitors tend to have a higher purchase rate than average (and when compared to returning visitors). Users also tend to have a higher-than-average purchase rate in November, October, and September.

So far, we've examined each feature individually and compared against `Revenue`. It would also be useful for us to compare features against each other to identify any collinearity. A scatterplot matrix of our numeric features (see Appendix, Figure A1) reveals that corresponding count-and-duration feature pairs lie almost on a line, and `BounceRates` and `ExitRates` show a strong positive linear relationship. Because of the collinearity between some of our features, we'll consider regularization in our models to handle redundancy.

## Data Preparation for Modeling

Before fitting any models, we need to encode categorical variables, split the data into training and test sets, and standardize numeric features. We convert `Revenue` to a factor, encode `Month` and `VisitorType` as dummy variables, and treat the categorical features with numeric labels (`OperatingSystems`, `Browser`, `Region`, `TrafficType`) as factors to be dummy-encoded as well. We also apply a log-transform to the skewed numeric features (excluding `BounceRates`, `ExitRates`, and `SpecialDay` which are bounded or discrete) to stabilize model training. We split the dataset $70/30$ into training and test sets. Full data preparation code is provided in the Appendix.

```{r data-prep}
library(caret)
library(MASS)
library(class)
library(e1071)
library(cluster)

# set seed
set.seed(432)

shoppers_model <- shoppers
shoppers_model$Revenue <- factor(shoppers_model$Revenue,
                                 levels = c(FALSE, TRUE),
                                 labels = c("No", "Yes"))
factor_cols <- c("Month", "VisitorType", "Weekend",
                 "OperatingSystems", "Browser", "Region", "TrafficType")
for (col in factor_cols) {
  shoppers_model[[col]] <- factor(shoppers_model[[col]])
}

dummy_formula <- dummyVars(Revenue ~ ., data = shoppers_model, fullRank = TRUE)
shoppers_encoded <- as.data.frame(predict(dummy_formula, newdata = shoppers_model))
shoppers_encoded$Revenue <- shoppers_model$Revenue

train_idx <- createDataPartition(shoppers_encoded$Revenue, p = 0.7, list = FALSE)
train_data <- shoppers_encoded[train_idx, ]
test_data <- shoppers_encoded[-train_idx, ]

train_x <- train_data[, names(train_data) != "Revenue"]
train_y <- train_data$Revenue
test_x <- test_data[, names(test_data) != "Revenue"]
test_y <- test_data$Revenue

nzv_cols <- nearZeroVar(train_x, names = TRUE)
if (length(nzv_cols) > 0) {
  train_x <- train_x[, !names(train_x) %in% nzv_cols]
  test_x <- test_x[, !names(test_x) %in% nzv_cols]
}

log_features <- c("Administrative", "Administrative_Duration",
                  "Informational", "Informational_Duration",
                  "ProductRelated", "ProductRelated_Duration",
                  "PageValues")
train_x_log <- train_x
test_x_log  <- test_x
for (col in log_features) {
  if (col %in% names(train_x_log)) {
    train_x_log[[col]] <- log1p(train_x_log[[col]])
    test_x_log[[col]]  <- log1p(test_x_log[[col]])
  }
}

preproc <- preProcess(train_x, method = c("center", "scale"))
train_x_scaled <- predict(preproc, train_x)
test_x_scaled <- predict(preproc, test_x)
```

## Logistic Regression

We fit four logistic regression variants: a default logistic regression without any penalty as baseline, Lasso ($L_1$), Ridge ($L_2$), and Elastic Net (a mix of both). To account for our skewness, we use our log-transformed features rather than their un-transformed counterparts. The penalized variants are motivated by the multicollinearity we observed between count/duration pairs and between `BounceRates` and `ExitRates`. For each penalized variant we use `cv.glmnet` with 10-fold cross-validation to pick $\lambda$, and for Elastic Net we additionally grid-search over $\alpha$.

```{r logistic-setup}
library(pROC)

report_metrics <- function(name, prob, truth) {
  pred <- factor(ifelse(prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
  cm   <- confusionMatrix(pred, truth, positive = "Yes")
  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1   <- ifelse(is.na(prec) || (sens + prec) == 0, 0,
                 2 * sens * prec / (sens + prec))
  auc_val <- as.numeric(auc(roc(truth, prob, levels = c("No", "Yes"), quiet = TRUE)))
  data.frame(Model = name, Accuracy = round(cm$overall["Accuracy"], 4),
             Sensitivity = round(sens, 4), Specificity = round(cm$byClass["Specificity"], 4),
             F1 = round(f1, 4), AUC = round(auc_val, 4), row.names = NULL)
}
```

### Baseline (Unpenalized) Logistic Regression

We start with a standard logistic regression on all encoded predictors as a benchmark.

```{r logistic-baseline}
set.seed(432)
log_train_df <- data.frame(Revenue = train_y, train_x_log)
log_test_df  <- data.frame(Revenue = test_y,  test_x_log)

log_baseline <- glm(Revenue ~ ., data = log_train_df, family = binomial())
log_baseline_prob <- predict(log_baseline, newdata = log_test_df, type = "response")
```

### Lasso-Penalized Logistic Regression ($L_1$)

Adding a Lasso penalty shrinks coefficients toward zero, performing automatic feature selection. We use `cv.glmnet` with 10-fold CV to pick the $\lambda$ that minimizes cross-validated deviance.

```{r logistic-lasso}
set.seed(432)
library(glmnet)

x_train_mat <- as.matrix(train_x_log)
x_test_mat  <- as.matrix(test_x_log)

lasso_cv <- cv.glmnet(x = x_train_mat, y = train_y,
                      family = "binomial", alpha = 1, nfolds = 10)
log_lasso_prob <- as.numeric(predict(lasso_cv, newx = x_test_mat,
                                     s = "lambda.min", type = "response"))
```

### Ridge-Penalized Logistic Regression ($L_2$)

Ridge shrinks all coefficients toward zero but never to exactly zero, helping handle collinearity between our features.

```{r logistic-ridge}
set.seed(432)

ridge_cv <- cv.glmnet(x = x_train_mat, y = train_y,
                      family = "binomial", alpha = 0, nfolds = 10)
log_ridge_prob <- as.numeric(predict(ridge_cv, newx = x_test_mat,
                                     s = "lambda.min", type = "response"))
```

### Elastic Net Logistic Regression

Elastic Net blends $L_1$ and $L_2$ via a mixing parameter $\alpha \in (0, 1)$. We grid-search over $\alpha$ and pick the combination with the lowest cross-validated error.

```{r logistic-elastic}
set.seed(432)

alpha_grid <- seq(0.1, 0.9, by = 0.1)
elastic_cv_results <- lapply(alpha_grid, function(a) {
  set.seed(432)
  cv.glmnet(x = x_train_mat, y = train_y,
            family = "binomial", alpha = a, nfolds = 10)
})

cv_min_errors  <- sapply(elastic_cv_results, function(fit) min(fit$cvm))
best_alpha_idx <- which.min(cv_min_errors)
best_alpha     <- alpha_grid[best_alpha_idx]
elastic_cv     <- elastic_cv_results[[best_alpha_idx]]
log_elastic_prob <- as.numeric(predict(elastic_cv, newx = x_test_mat,
                                       s = "lambda.min", type = "response"))
```

### Logistic Regression Comparison

```{r logistic-comparison}
logreg_comparison <- rbind(
  report_metrics("Baseline", log_baseline_prob, test_y),
  report_metrics("Lasso", log_lasso_prob, test_y),
  report_metrics("Ridge", log_ridge_prob, test_y),
  report_metrics("Elastic Net", log_elastic_prob, test_y)
)
knitr::kable(logreg_comparison, caption = "Logistic Regression Variants: Test Set Performance")

lasso_pred <- factor(ifelse(log_lasso_prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
lasso_cm <- confusionMatrix(lasso_pred, test_y, positive = "Yes")
```

The Lasso and Elastic Net models perform similarly and slightly outperform the unpenalized baseline, confirming that regularization helps in the presence of collinear features. Ridge performs slightly worse, likely because it retains all features rather than zeroing out uninformative ones. Overall, the Lasso-penalized model offers the best balance of simplicity and performance among the logistic regression variants.

## Unsupervised Learning

### Principal Component Analysis (PCA)

We apply PCA to the 10 original numeric features (before dummy encoding) to understand the variance structure and to provide reduced-dimension inputs for later models.

```{r pca, fig.width=9, fig.height=3.5}
train_numeric <- train_data %>% dplyr::select(all_of(numeric_features))
test_numeric <- test_data %>% dplyr::select(all_of(numeric_features))

pca_preproc <- preProcess(train_numeric, method = c("center", "scale"))
train_numeric_scaled <- predict(pca_preproc, train_numeric)
test_numeric_scaled <- predict(pca_preproc, test_numeric)

pca_fit <- prcomp(train_numeric_scaled, center = FALSE, scale. = FALSE)

var_explained <- summary(pca_fit)$importance
par(mfrow = c(1, 2))
plot(var_explained[2, ], type = "b", pch = 19, col = "steelblue",
     xlab = "Principal Component", ylab = "Proportion of Variance",
     main = "Scree Plot")
plot(var_explained[3, ], type = "b", pch = 19, col = "darkred",
     xlab = "Principal Component", ylab = "Cumulative Proportion",
     main = "Cumulative Variance Explained")
abline(h = 0.9, lty = 2, col = "gray40")
par(mfrow = c(1, 1))
```

```{r pca-biplot, fig.width=7, fig.height=4}
train_pca_scores <- as.data.frame(pca_fit$x[, 1:2])
train_pca_scores$Revenue <- train_y

ggplot(train_pca_scores, aes(x = PC1, y = PC2, color = Revenue)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "PCA: First Two Principal Components Colored by Revenue",
       x = paste0("PC1 (", round(var_explained[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2, 2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "gray60", "Yes" = "tomato"))
```

The PCA biplot shows that the two classes overlap substantially in the first two principal components, indicating that separating purchasers from non-purchasers requires more than two linear directions. However, `Revenue = Yes` sessions tend to have a wider spread, particularly along PC1, suggesting that high-engagement sessions (which load onto the page count and duration variables) are more likely to result in purchases.

### K-Means Clustering

We cluster the scaled numeric features to discover natural session groupings. We use the silhouette method to choose the number of clusters.

```{r kmeans-silhouette, fig.width=7, fig.height=5}
set.seed(432)
sample_idx <- sample(nrow(train_numeric_scaled), 3000)
train_sample <- train_numeric_scaled[sample_idx, ]

sil_widths <- sapply(2:8, function(k) {
  km <- kmeans(train_sample, centers = k, nstart = 25, iter.max = 50)
  ss <- silhouette(km$cluster, dist(train_sample))
  mean(ss[, 3])
})

plot(2:8, sil_widths, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (k)", ylab = "Average Silhouette Width",
     main = "Silhouette Analysis for K-Means")
best_k <- which.max(sil_widths) + 1
```

```{r kmeans-fit, fig.width=9, fig.height=5}
km_fit <- kmeans(train_numeric_scaled, centers = best_k, nstart = 25, iter.max = 100)

train_pca_all <- as.data.frame(pca_fit$x[, 1:2])
train_pca_all$Cluster <- factor(km_fit$cluster)
train_pca_all$Revenue <- train_y

p1 <- ggplot(train_pca_all, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "K-Means Clusters in PCA Space") +
  theme_minimal()

# Per-cluster sample size and purchase rate (for the bar plot below)
cluster_n <- table(train_pca_all$Cluster)
cluster_rates <- tapply(train_pca_all$Revenue == "Yes",
                        train_pca_all$Cluster, mean)
cluster_revenue <- data.frame(
  Cluster = factor(names(cluster_n), levels = names(cluster_n)),
  n = as.numeric(cluster_n),
  purchase_rate = as.numeric(cluster_rates)
)

p2 <- ggplot(cluster_revenue, aes(x = Cluster, y = purchase_rate, fill = Cluster)) +
  geom_col() +
  geom_hline(yintercept = mean(train_y == "Yes"), linetype = "dashed") +
  labs(title = "Purchase Rate by Cluster",
       subtitle = "Dashed line = overall training rate",
       y = "Purchase Rate") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

```{r kmeans-profiles}
# Mean of each numeric feature within each cluster (original scale, not scaled).
# aggregate() returns one row per cluster with feature-mean columns.
cluster_profiles <- aggregate(
  train_data[, numeric_features],
  by = list(Cluster = factor(km_fit$cluster)),
  FUN = mean
)

cluster_profiles_long <- cluster_profiles %>%
  pivot_longer(-Cluster, names_to = "Feature", values_to = "Mean") %>%
  pivot_wider(names_from = Cluster, values_from = Mean, names_prefix = "Cluster ")
```

Cluster centroid profiles and a hierarchical clustering dendrogram corroborating these groupings are provided in the Appendix (Table A1 and Figure A2).

## Supervised Learning

We now turn to classification. For all supervised models, we evaluate using 5-fold stratified cross-validation on the training set. Because the classes are imbalanced ($\sim 85\%$ non-purchase), we report not just accuracy but also sensitivity (recall for purchases), specificity, and AUC. We collect results into a summary table at the end.

```{r cv-setup}
# trainControl is used by the caret-based models below (LDA, KNN).
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

results <- list()

results[["Lasso LogReg"]] <- list(
  cm = lasso_cm,
  probs = data.frame(No = 1 - log_lasso_prob, Yes = log_lasso_prob)
)
```

### Discriminant Analysis: Choosing Between LDA, QDA, and Naive Bayes

Among generative classifiers, we consider three options: Linear Discriminant Analysis (LDA), Quadratic Discriminant Analysis (QDA), and Naive Bayes. We select **LDA** for the following reasons:

- **QDA** requires estimating a separate covariance matrix for each class. With our high-dimensional dummy-encoded feature space, the within-class covariance matrices become singular or near-singular, forcing us to restrict QDA to a small PCA subspace and lose information from the categorical predictors.
- **Naive Bayes** assumes conditional independence among features given the class — a strong assumption clearly violated here (e.g., `BounceRates` and `ExitRates` are highly correlated, as are page counts and durations). While Naive Bayes can still produce reasonable rankings, its probability estimates are poorly calibrated under dependence, and it typically underperforms LDA when the shared-covariance assumption is approximately satisfied.
- **LDA** assumes multivariate normality with a shared covariance matrix across classes. While normality is imperfect for our skewed features, the shared-covariance assumption is more defensible than independence, and LDA can operate on the full encoded feature set without dimensionality issues. It also provides a natural projection onto the discriminant axis, aiding interpretation.

### Linear Discriminant Analysis (LDA)

LDA projects the data onto the single linear discriminant axis that maximally separates the two classes. Given the collinearity we observed earlier, we should expect LDA to perform comparably to logistic regression.

```{r lda-fit}
lda_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "lda",
  trControl = ctrl,
  metric = "ROC"
)

lda_pred <- predict(lda_model, test_x_scaled)
lda_prob <- predict(lda_model, test_x_scaled, type = "prob")

lda_cm <- confusionMatrix(lda_pred, test_y, positive = "Yes")

results[["LDA"]] <- list(
  model = lda_model,
  cm = lda_cm,
  probs = lda_prob
)
```

### K-Nearest Neighbors (KNN)

KNN classifies each observation by a majority vote of its $k$ nearest neighbors. It is a non-parametric, lazy-learning method — no model is fitted at training time; all computation happens at prediction time. This makes KNN impractical for real-time session scoring (as noted in the original paper), but it provides a useful non-parametric reference. We tune $k$ via cross-validation. Given the curse of dimensionality, we also try KNN on PCA-reduced features.

```{r knn-fit}
knn_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pred <- predict(knn_model, test_x_scaled)
knn_prob <- predict(knn_model, test_x_scaled, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_y, positive = "Yes")

results[["KNN"]] <- list(
  model = knn_model,
  cm = knn_cm,
  probs = knn_prob
)
```

```{r knn-pca, fig.width=7, fig.height=5}
train_pca_full <- as.data.frame(pca_fit$x[, 1:7])
test_pca_full <- as.data.frame(predict(pca_fit, test_numeric_scaled)[, 1:7])

knn_pca_model <- train(
  x = train_pca_full,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pca_pred <- predict(knn_pca_model, test_pca_full)
knn_pca_prob <- predict(knn_pca_model, test_pca_full, type = "prob")

knn_pca_cm <- confusionMatrix(knn_pca_pred, test_y, positive = "Yes")

results[["KNN (PCA)"]] <- list(
  model = knn_pca_model,
  cm = knn_pca_cm,
  probs = knn_pca_prob
)
```

## Tree Methods: Bagging, Boosting, and Random Forest

We now consider tree-based classifiers. We use the log-transformed features prepared earlier.

### Random Forest

We fit a random forest with 1,000 trees and tune `mtry` (the number of features considered at each split) using out-of-bag (OOB) error. OOB error is the natural evaluation metric for bagged models: each tree is trained on a bootstrap sample, and the observations left out of that sample provide a built-in validation set without needing a separate cross-validation loop.

```{r rf-fit, results='hide'}
library(randomForest)

ctrl_oob <- trainControl(method = "oob", classProbs = TRUE)
rf <- train(
  x = train_x_log, y = train_y, method = "rf",
  trControl = ctrl_oob,
  tuneGrid = expand.grid(mtry = c(5, 10, 15, 20, ncol(train_x_log))),
  ntree = 1000, importance = TRUE
)
```

```{r rf-results}
knitr::kable(rf$results[, c("mtry", "Accuracy", "Kappa")], digits = 4,
             caption = "Random Forest: OOB Performance by mtry")
```

The optimal `mtry` is `r rf$bestTune$mtry`, achieving an OOB accuracy of `r round(rf$results$Accuracy[which.min(rf$results$Accuracy == max(rf$results$Accuracy))], 4)`. Note that the bagging model (`mtry` = `r ncol(train_x_log)`, using all predictors) performs worse than the random forest, confirming that decorrelating the trees by restricting the feature set at each split improves ensemble performance.

```{r rf-test, fig.height=4, fig.width=10}
rf_pred <- predict(rf, newdata = test_x_log)
rf_cm <- confusionMatrix(rf_pred, test_y, positive = "Yes")

varImpPlot(rf$finalModel, main = "Variable Importance for Online Shoppers")
```

`PageValues` is by far the most important variable by both mean decrease in accuracy and mean decrease in Gini, consistent with its role as a direct proxy for purchase intent. The test accuracy is `r round(rf_cm$overall["Accuracy"], 4)`.

### Gradient Boosting (GBM)

Unlike random forests, boosting builds trees sequentially, with each tree correcting the errors of the previous ensemble. We tune the number of trees, interaction depth, and learning rate via 5-fold cross-validation on accuracy. We use accuracy rather than OOB error because boosting does not employ bootstrap aggregation.

```{r gbm-fit, results='hide'}
gbm_grid <- expand.grid(
  n.trees = c(500, 1000),
  interaction.depth = c(3, 4),
  shrinkage = c(0.01, 0.1),
  n.minobsinnode = 10
)

ctrl_accuracy <- trainControl(
  method = "cv", number = 5,
  classProbs = TRUE, summaryFunction = defaultSummary
)

combined <- as.data.frame(train_x_log)
combined$Revenue <- train_y

gbm_fit <- train(
  Revenue ~ ., data = combined, method = "gbm",
  trControl = ctrl_accuracy, tuneGrid = gbm_grid,
  metric = "Accuracy", verbose = FALSE
)
```

```{r gbm-results}
best_gbm <- gbm_fit$bestTune
cat("Best GBM: n.trees =", best_gbm$n.trees,
    ", depth =", best_gbm$interaction.depth,
    ", shrinkage =", best_gbm$shrinkage, "\n")

gbm_imp <- summary(gbm_fit$finalModel, plotit = FALSE)
knitr::kable(head(gbm_imp, 10), digits = 2,
             caption = "GBM: Top 10 Variables by Relative Influence")
```

Again, `PageValues` dominates the relative influence rankings. The optimal model uses `r best_gbm$n.trees` trees with interaction depth `r best_gbm$interaction.depth` and a learning rate of `r best_gbm$shrinkage`.

```{r gbm-test}
gbm_pred <- predict(gbm_fit, newdata = test_x_log)
gbm_cm <- confusionMatrix(gbm_pred, test_y, positive = "Yes")
```

The GBM test accuracy is `r round(gbm_cm$overall["Accuracy"], 4)`, slightly higher than the random forest.

```{r tree-store-results}
results[["Random Forest"]] <- list(
  cm = rf_cm,
  probs = predict(rf, newdata = test_x_log, type = "prob")
)

results[["GBM"]] <- list(
  cm = gbm_cm,
  probs = predict(gbm_fit, newdata = test_x_log, type = "prob")
)
```

## Model Comparison

We now compare all models on key metrics computed from their test set predictions.

```{r model-comparison, fig.width=9, fig.height=5}
library(pROC)

comparison <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  F1 = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (name in names(results)) {
  cm <- results[[name]]$cm
  probs <- results[[name]]$probs

  roc_obj <- roc(test_y, probs$Yes, levels = c("No", "Yes"), quiet = TRUE)

  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1 <- ifelse(is.na(prec) | (sens + prec) == 0, 0, 2 * sens * prec / (sens + prec))

  comparison <- rbind(comparison, data.frame(
    Model = name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(sens, 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1 = round(f1, 4),
    AUC = round(auc(roc_obj), 4),
    stringsAsFactors = FALSE
  ))
}

rownames(comparison) <- NULL
knitr::kable(comparison, caption = "Model Comparison on Test Set")
```

```{r roc-curves, fig.width=8, fig.height=4}
roc_list <- list()
for (name in names(results)) {
  roc_list[[name]] <- roc(test_y, results[[name]]$probs$Yes,
                          levels = c("No", "Yes"), quiet = TRUE)
}

plot(roc_list[[1]], col = 1, main = "ROC Curves: Model Comparison")
for (i in 2:length(roc_list)) {
  lines(roc_list[[i]], col = i)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=",
                       sapply(roc_list, function(r) round(auc(r), 3)), ")"),
       col = 1:length(roc_list), lwd = 2, cex = 0.8)
```

The comparison table and ROC curves reveal several patterns. The tree-based ensemble methods (GBM and Random Forest) achieve the highest accuracy and AUC, benefiting from their ability to capture nonlinear interactions and handle the mixed feature space. LDA performs well among the simpler methods due to the strong linear signal in `PageValues` and `ExitRates`. All models show relatively low sensitivity, a direct consequence of the $\sim 85\%$ class imbalance: with a default 0.5 threshold, classifiers are conservative and rarely predict a purchase unless the evidence is overwhelming. KNN suffers from the curse of dimensionality, though PCA-reduced features partially mitigate this. Across all methods, `PageValues` consistently emerges as the dominant predictor of purchase intent.

## Conclusion

Among the models evaluated, Gradient Boosted Trees achieved the best overall test performance, followed closely by Random Forest. Both ensemble methods substantially outperformed the simpler classifiers (LDA, KNN) in accuracy and AUC, owing to their capacity for modeling nonlinear feature interactions and their robustness to the mixed numeric-categorical feature space. We recommend GBM as the primary classifier for this task, with Random Forest as a strong alternative that offers more interpretable variable importance measures. Regardless of method, the class imbalance remains the central challenge: all models exhibit high specificity but limited sensitivity, indicating that threshold tuning or oversampling strategies would be necessary for deployment in a real-time purchase-intent flagging system.

### Future Work

From the tree models, the variable that is most important across all methods — both in mean decrease in accuracy and mean decrease in Gini — is `PageValues`. Conversely, the least important variables include several `TrafficType` levels, `Weekend`, `Month`, operating system indicators, and some `Region` levels. This suggests that temporal and geographic attributes do not contribute substantially to predicting revenue, nor does the operating system used by a potential customer. These findings are largely unsurprising, with the notable exception of `Month`: one might expect seasonality to play a significant role in purchasing behavior, given that different months bring different occasions and shopping surges. The relatively low importance of `Month` warrants further investigation — for instance, whether its effect is absorbed by `SpecialDay` or whether a more granular temporal encoding (e.g., week-of-year) would reveal stronger seasonal signal.

\newpage

## Appendix

### Data Preparation Code

```{r appendix-data-prep, echo=TRUE, eval=FALSE}
library(caret); library(MASS); library(class); library(e1071); library(cluster)
set.seed(432)

# Convert Revenue and categorical features to factors
shoppers_model <- shoppers
shoppers_model$Revenue <- factor(shoppers_model$Revenue,
                                 levels = c(FALSE, TRUE), labels = c("No", "Yes"))
factor_cols <- c("Month", "VisitorType", "Weekend",
                 "OperatingSystems", "Browser", "Region", "TrafficType")
for (col in factor_cols) shoppers_model[[col]] <- factor(shoppers_model[[col]])

# One-hot encode categoricals
dummy_formula <- dummyVars(Revenue ~ ., data = shoppers_model, fullRank = TRUE)
shoppers_encoded <- as.data.frame(predict(dummy_formula, newdata = shoppers_model))
shoppers_encoded$Revenue <- shoppers_model$Revenue

# Stratified 70/30 train/test split
train_idx <- createDataPartition(shoppers_encoded$Revenue, p = 0.7, list = FALSE)
train_data <- shoppers_encoded[train_idx, ]; test_data <- shoppers_encoded[-train_idx, ]
train_x <- train_data[, names(train_data) != "Revenue"]; train_y <- train_data$Revenue
test_x <- test_data[, names(test_data) != "Revenue"]; test_y <- test_data$Revenue

# Remove near-zero variance columns
nzv_cols <- nearZeroVar(train_x, names = TRUE)
if (length(nzv_cols) > 0) {
  train_x <- train_x[, !names(train_x) %in% nzv_cols]
  test_x <- test_x[, !names(test_x) %in% nzv_cols]
}

# Log-transform skewed numeric features (skip bounded/discrete ones)
log_features <- c("Administrative", "Administrative_Duration",
                  "Informational", "Informational_Duration",
                  "ProductRelated", "ProductRelated_Duration", "PageValues")
train_x_log <- train_x; test_x_log <- test_x
for (col in log_features) {
  if (col %in% names(train_x_log)) {
    train_x_log[[col]] <- log1p(train_x_log[[col]])
    test_x_log[[col]]  <- log1p(test_x_log[[col]])
  }
}

# Center and scale for distance-based methods
preproc <- preProcess(train_x, method = c("center", "scale"))
train_x_scaled <- predict(preproc, train_x)
test_x_scaled <- predict(preproc, test_x)
```

### Figure A1: Scatterplot Matrix

```{r scatterplot-matrix, fig.width=10, fig.height=10}
shoppers_numeric <- subset(shoppers, select = numeric_features)
pairs(shoppers_numeric,
      main = "Scatterplot Matrix for Numeric Features",
      col = "navy",
      pch = 20)
```

### Table A1: Cluster Centroids

```{r appendix-cluster-profiles}
knitr::kable(cluster_profiles_long, digits = 2, caption = "Cluster Centroids (Original Scale)")
```

### Figure A2: Hierarchical Clustering Dendrogram

```{r appendix-hierarchical, fig.width=9, fig.height=5}
hc_dist <- dist(train_sample)
hc_fit <- hclust(hc_dist, method = "ward.D2")

plot(hc_fit, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method, n=3000 sample)",
     xlab = "", sub = "")
rect.hclust(hc_fit, k = best_k, border = "steelblue")
```

### Full Code

All code for this project is available at: https://github.com/VelupallyG/432-online-shoppers-intention