---
title: "STAT432 Final Project"
output: pdf_document
date: "2026-04-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction and Contextualization

"Online Shoppers Purchasing Intention Dataset" provided by UCI Machine Learning Repository [here](https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset).

The dataset contains $12{,}330$ observations of online shopping sessions, with $17$ input variables and one categorical output variable, `Revenue`, indicating whether the session ended in a purchase. The classes are imbalanced, with only about $15\%$ of sessions resulted in a purchase. There is no missing data.

We divide the $17$ predictors into four conceptual groups:

- **Page Interaction Features**: the number of pages of each type visited in the session and the total time spent on them: `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, and `ProductRelated_Duration`. These are derived from URL information and updated in real time as the user moves between pages.

- **Google Analytics Metrics** — engagement-quality metrics from Google Analytics: `BounceRates` (the percentage of visitors who enter the site from a page and leave without triggering any further requests), `ExitRates` (the percentage of pageviews on a page that were the last in the session), and `PageValues` (the average value of a page that a user visited before completing a transaction).

- **Session Context Features** — time- and seasonality-related variables: `SpecialDay`, `Month`, and `Weekend`. The `SpecialDay` value reflects how close the visit is to a holiday such as Valentine's Day or Mother's Day; for Valentine's Day, for instance, it takes a nonzero value between February 2 and February 12 and peaks at $1$ on February 8.

- **User Technical Profile** — session-level user attributes: `OperatingSystems`, `Browser`, `Region`, `TrafficType`, and `VisitorType` (whether the user is a new or returning visitor).

Our goal is to build classification models that can effectively determine whether or not an online shopper will make a purchase based on the feature variables above.

## Data Preprocessing and Visualization

We'll start by loading in our data.

```{r load-data, message=FALSE}
library(tidyverse)

shoppers <- read_csv("online_shoppers_intention.csv")
glimpse(shoppers)
```

Before fitting any models, we want to inspect the distribution of each input variable individually. This helps us visually identify features that may be skewed or that may follow a distribution that we can take advantage of. We separate variables into numeric and categorical groups, plotting histograms for numeric features and bar plots for categorical features.

```{r univariate-numeric, fig.width=9, fig.height=7}
numeric_features <- c(
  "Administrative", "Administrative_Duration",
  "Informational", "Informational_Duration",
  "ProductRelated", "ProductRelated_Duration",
  "BounceRates", "ExitRates", "PageValues", "SpecialDay"
)

shoppers %>%
  select(all_of(numeric_features)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  labs(title = "Distributions of Numeric Features", x = NULL, y = "Count") +
  theme_minimal()
```

After plotting, we notice that each of our numeric features is mostly centered near $0$ and have skew to the right. We may want to perform some log transformation of our features to handle the skew. Notably, `SpecialDay` contains discrete values which may require special handling.

```{r univariate-categorical, fig.width=9, fig.height=7}
categorical_features <- c(
  "Month", "OperatingSystems", "Browser",
  "Region", "TrafficType", "VisitorType", "Weekend"
)

shoppers %>%
  select(all_of(categorical_features)) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_bar(fill = "steelblue") +
  facet_wrap(~ feature, scales = "free", ncol = 3) +
  labs(title = "Distributions of Categorical Features", x = NULL, y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Several of the categorical distributions are heavily imbalanced. `Browser`, `OperatingSystems`, and `TrafficType` are dominated by one or two levels with a long tail of rare ones, which suggests we may want to lump rare levels together (e.g. via `forcats::fct_lump`) before modeling so that the rare levels don't introduce instability. `VisitorType` is mostly returning visitors with a smaller share of new visitors, and `Weekend` is mostly `FALSE`. `Month` shows a seasonal pattern with clear peaks (notably in May and November). Whether features like `OperatingSystems` or `Browser` carry real predictive signal is an open question — we'll get a better sense of that in the bivariate plots below.

### Bivariate Distributions vs. Revenue

Univariate plots tell us how each feature is distributed in isolation, but for a classification problem the more useful question is: *does this feature differ between sessions that ended in a purchase and those that did not?* If the conditional distributions of a feature given `Revenue = TRUE` and `Revenue = FALSE` look similar, that feature carries little signal; if they look different, it likely matters. We use boxplots for the numeric features (comparing distributions by `Revenue`) and stacked proportion bars for the categorical features (showing the proportion of purchases within each level relative to the $\sim 15\%$ overall purchase rate).

```{r bivariate-numeric, fig.width=9, fig.height=7}
shoppers %>%
  select(all_of(numeric_features), Revenue) %>%
  pivot_longer(-Revenue, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = Revenue, y = value, fill = Revenue)) +
  geom_boxplot(outlier.size = 0.4, alpha = 0.8) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  labs(title = "Numeric Features by Revenue", x = "Revenue", y = NULL) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r bivariate-categorical, fig.width=9, fig.height=7}
shoppers %>%
  select(all_of(categorical_features), Revenue) %>%
  mutate(across(-Revenue, as.character)) %>%
  pivot_longer(-Revenue, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = Revenue)) +
  geom_bar(position = "fill") +
  geom_hline(yintercept = mean(shoppers$Revenue), linetype = "dashed", color = "black") +
  facet_wrap(~ feature, scales = "free_x", ncol = 3) +
  labs(title = "Purchase Proportion by Categorical Feature",
       subtitle = "Dashed line marks the overall purchase rate (~15%)",
       x = NULL, y = "Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Scatterplot Matrix

So far we've looked at each predictor on its own and against the target, but we haven't yet examined how the predictors relate to one another. Several of our numeric features are likely to be strongly related by construction — for example, the count of pages of a given type and the total time spent on those pages should rise together, and `BounceRates` and `ExitRates` measure closely related notions of leaving the site. Highly collinear predictors are a problem for linear models like logistic regression (unstable coefficients, inflated standard errors) and motivate the use of regularization methods like Lasso or Ridge. A scatterplot matrix on the numeric features lets us see these pairwise relationships directly, including any nonlinear patterns that a single correlation coefficient would miss.

```{r scatterplot-matrix, fig.width=10, fig.height=10}
shoppers_numeric <- subset(shoppers, select = numeric_features)
pairs(shoppers_numeric,
      main = "Scatterplot Matrix for Numeric Features",
      col = "navy",
      pch = 20)
```

A few clear patterns stand out from the matrix. Most obviously, each count-and-duration pair lies almost exactly on a line: `Administrative` with `Administrative_Duration`, `Informational` with `Informational_Duration`, and `ProductRelated` with `ProductRelated_Duration`. This is unsurprising — visiting more pages of a given type produces more total time spent on those pages — but it means each pair carries largely the same information, so we should expect collinearity issues with linear models if we include both. `BounceRates` and `ExitRates` also show a strong positive linear relationship, which makes sense given that both measure forms of leaving the site; we will likely want to include only one of them, or rely on regularization to handle the redundancy. Beyond these tight pairs, the relationships are weaker and harder to read because of the heavy right-skew in most features (most points are clustered near the origin with a long tail of outliers), reinforcing the case for log-transforming these variables before fitting linear models. `PageValues` and `SpecialDay` do not appear to be strongly related to any of the other features.

## Data Preparation for Modeling

Before fitting any models, we need to encode categorical variables, split the data into training and test sets, and standardize numeric features. We convert `Revenue` to a factor, encode `Month` and `VisitorType` as dummy variables, and treat the integer-coded nominal features (`OperatingSystems`, `Browser`, `Region`, `TrafficType`) as factors to be dummy-encoded as well.

```{r data-prep, message=FALSE}
library(caret)
library(MASS)
library(class)
library(e1071)
library(cluster)

set.seed(432)

shoppers_model <- shoppers %>%
  mutate(
    Revenue = factor(Revenue, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    Month = factor(Month),
    VisitorType = factor(VisitorType),
    Weekend = factor(Weekend),
    OperatingSystems = factor(OperatingSystems),
    Browser = factor(Browser),
    Region = factor(Region),
    TrafficType = factor(TrafficType)
  )

dummy_formula <- dummyVars(Revenue ~ ., data = shoppers_model, fullRank = TRUE)
shoppers_encoded <- predict(dummy_formula, newdata = shoppers_model) %>% as.data.frame()
shoppers_encoded$Revenue <- shoppers_model$Revenue

train_idx <- createDataPartition(shoppers_encoded$Revenue, p = 0.7, list = FALSE)
train_data <- shoppers_encoded[train_idx, ]
test_data <- shoppers_encoded[-train_idx, ]

train_x <- train_data %>% dplyr::select(-Revenue)
train_y <- train_data$Revenue
test_x <- test_data %>% dplyr::select(-Revenue)
test_y <- test_data$Revenue

nzv_cols <- nearZeroVar(train_x, names = TRUE)
if (length(nzv_cols) > 0) {
  cat("Removing near-zero variance columns:", paste(nzv_cols, collapse = ", "), "\n")
  train_x <- train_x[, !names(train_x) %in% nzv_cols]
  test_x <- test_x[, !names(test_x) %in% nzv_cols]
}

preproc <- preProcess(train_x, method = c("center", "scale"))
train_x_scaled <- predict(preproc, train_x)
test_x_scaled <- predict(preproc, test_x)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
cat("Training class distribution:\n")
table(train_y) %>% print()
cat("Number of features after encoding:", ncol(train_x), "\n")
```

## Unsupervised Learning

### Principal Component Analysis (PCA)

We apply PCA to the 10 original numeric features (before dummy encoding) to understand the variance structure and to provide reduced-dimension inputs for later models.

```{r pca, fig.width=9, fig.height=5}
train_numeric <- train_data %>% dplyr::select(all_of(numeric_features))
test_numeric <- test_data %>% dplyr::select(all_of(numeric_features))

pca_preproc <- preProcess(train_numeric, method = c("center", "scale"))
train_numeric_scaled <- predict(pca_preproc, train_numeric)
test_numeric_scaled <- predict(pca_preproc, test_numeric)

pca_fit <- prcomp(train_numeric_scaled, center = FALSE, scale. = FALSE)

var_explained <- summary(pca_fit)$importance
par(mfrow = c(1, 2))
plot(var_explained[2, ], type = "b", pch = 19, col = "steelblue",
     xlab = "Principal Component", ylab = "Proportion of Variance",
     main = "Scree Plot")
plot(var_explained[3, ], type = "b", pch = 19, col = "darkred",
     xlab = "Principal Component", ylab = "Cumulative Proportion",
     main = "Cumulative Variance Explained")
abline(h = 0.9, lty = 2, col = "gray40")
par(mfrow = c(1, 1))

cat("Variance explained by first 5 PCs:", round(var_explained[3, 5], 3), "\n")
cat("Variance explained by first 7 PCs:", round(var_explained[3, 7], 3), "\n")
```

```{r pca-biplot, fig.width=8, fig.height=6}
train_pca_scores <- as.data.frame(pca_fit$x[, 1:2])
train_pca_scores$Revenue <- train_y

ggplot(train_pca_scores, aes(x = PC1, y = PC2, color = Revenue)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "PCA: First Two Principal Components Colored by Revenue",
       x = paste0("PC1 (", round(var_explained[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2, 2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "gray60", "Yes" = "tomato"))
```

The PCA biplot shows that the two classes overlap substantially in the first two principal components, indicating that separating purchasers from non-purchasers requires more than two linear directions. However, `Revenue = Yes` sessions tend to have a wider spread, particularly along PC1, suggesting that high-engagement sessions (which load onto the page count and duration variables) are more likely to result in purchases.

### K-Means Clustering

We cluster the scaled numeric features to discover natural session groupings. We use the silhouette method to choose the number of clusters.

```{r kmeans-silhouette, fig.width=7, fig.height=5}
set.seed(432)
sample_idx <- sample(nrow(train_numeric_scaled), 3000)
train_sample <- train_numeric_scaled[sample_idx, ]

sil_widths <- sapply(2:8, function(k) {
  km <- kmeans(train_sample, centers = k, nstart = 25, iter.max = 50)
  ss <- silhouette(km$cluster, dist(train_sample))
  mean(ss[, 3])
})

plot(2:8, sil_widths, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (k)", ylab = "Average Silhouette Width",
     main = "Silhouette Analysis for K-Means")
best_k <- which.max(sil_widths) + 1
cat("Best k by silhouette:", best_k, "\n")
```

```{r kmeans-fit, fig.width=9, fig.height=5}
km_fit <- kmeans(train_numeric_scaled, centers = best_k, nstart = 25, iter.max = 100)

train_pca_all <- as.data.frame(pca_fit$x[, 1:2])
train_pca_all$Cluster <- factor(km_fit$cluster)
train_pca_all$Revenue <- train_y

p1 <- ggplot(train_pca_all, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "K-Means Clusters in PCA Space") +
  theme_minimal()

cluster_revenue <- train_pca_all %>%
  group_by(Cluster) %>%
  summarise(
    n = n(),
    purchase_rate = mean(Revenue == "Yes"),
    .groups = "drop"
  )

p2 <- ggplot(cluster_revenue, aes(x = Cluster, y = purchase_rate, fill = Cluster)) +
  geom_col() +
  geom_hline(yintercept = mean(train_y == "Yes"), linetype = "dashed") +
  labs(title = "Purchase Rate by Cluster",
       subtitle = "Dashed line = overall training rate",
       y = "Purchase Rate") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
print(cluster_revenue)
```

```{r kmeans-profiles}
cluster_profiles <- train_data %>%
  dplyr::select(all_of(numeric_features)) %>%
  mutate(Cluster = factor(km_fit$cluster)) %>%
  group_by(Cluster) %>%
  summarise(across(everything(), mean), .groups = "drop")

cluster_profiles_long <- cluster_profiles %>%
  pivot_longer(-Cluster, names_to = "Feature", values_to = "Mean") %>%
  pivot_wider(names_from = Cluster, values_from = Mean, names_prefix = "Cluster ")

knitr::kable(cluster_profiles_long, digits = 2, caption = "Cluster Centroids (Original Scale)")
```

### Hierarchical Clustering

We also apply hierarchical clustering (Ward's method) to a subsample and display a dendrogram to reason about the natural number of groups.

```{r hierarchical, fig.width=9, fig.height=5}
hc_dist <- dist(train_sample)
hc_fit <- hclust(hc_dist, method = "ward.D2")

plot(hc_fit, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method, n=3000 sample)",
     xlab = "", sub = "")
rect.hclust(hc_fit, k = best_k, border = "steelblue")
```

The dendrogram corroborates the K-means result: the data naturally splits into a small number of groups, broadly distinguishing low-engagement sessions from high-engagement ones.

## Supervised Learning

We now turn to classification. For all supervised models, we evaluate using 5-fold stratified cross-validation on the training set. Because the classes are imbalanced ($\sim 85\%$ non-purchase), we report not just accuracy but also sensitivity (recall for purchases), specificity, and AUC. We collect results into a summary table at the end.

```{r cv-setup}
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

results <- list()
```

### Linear Discriminant Analysis (LDA)

LDA assumes each class has a multivariate normal distribution with a shared covariance matrix. It projects the data onto the single linear discriminant axis that maximally separates the two classes. Given the collinearity we observed earlier, we should expect LDA to perform comparably to logistic regression.

```{r lda-fit}
lda_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "lda",
  trControl = ctrl,
  metric = "ROC"
)

lda_pred <- predict(lda_model, test_x_scaled)
lda_prob <- predict(lda_model, test_x_scaled, type = "prob")

lda_cm <- confusionMatrix(lda_pred, test_y, positive = "Yes")
cat("LDA Test Results:\n")
print(lda_cm)

results[["LDA"]] <- list(
  model = lda_model,
  cm = lda_cm,
  probs = lda_prob
)
```

### Quadratic Discriminant Analysis (QDA)

QDA relaxes LDA's assumption of equal covariance matrices, allowing each class to have its own covariance structure. This additional flexibility may help if the purchase and non-purchase sessions differ not just in their means but also in their variance patterns. However, QDA estimates more parameters and can be unstable with many features, especially when some are near-collinear. We use the dummy-encoded features but may need to apply PCA first if QDA fails on the full feature set.

```{r qda-fit, warning=FALSE}
train_pca_full <- as.data.frame(pca_fit$x[, 1:7])
test_pca_full <- as.data.frame(predict(pca_fit, test_numeric_scaled)[, 1:7])

qda_model <- train(
  x = train_pca_full,
  y = train_y,
  method = "qda",
  trControl = ctrl,
  metric = "ROC"
)

qda_pred <- predict(qda_model, test_pca_full)
qda_prob <- predict(qda_model, test_pca_full, type = "prob")

qda_cm <- confusionMatrix(qda_pred, test_y, positive = "Yes")
cat("QDA Test Results (on 7 PCs from numeric features):\n")
print(qda_cm)

results[["QDA"]] <- list(
  model = qda_model,
  cm = qda_cm,
  probs = qda_prob
)
```

### Naive Bayes

Naive Bayes assumes that features are conditionally independent given the class label. This is almost certainly violated in our data (e.g., `BounceRates` and `ExitRates` are strongly correlated), but the method can still perform surprisingly well in practice because it only needs to estimate the correct *decision boundary*, not the full joint distribution. We apply Naive Bayes to the full encoded feature set.

```{r nb-fit, warning=FALSE}
nb_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "naive_bayes",
  trControl = ctrl,
  metric = "ROC"
)

nb_pred <- predict(nb_model, test_x_scaled)
nb_prob <- predict(nb_model, test_x_scaled, type = "prob")

nb_cm <- confusionMatrix(nb_pred, test_y, positive = "Yes")
cat("Naive Bayes Test Results:\n")
print(nb_cm)

results[["Naive Bayes"]] <- list(
  model = nb_model,
  cm = nb_cm,
  probs = nb_prob
)
```

### K-Nearest Neighbors (KNN)

KNN classifies each observation by a majority vote of its $k$ nearest neighbors. It is a non-parametric, lazy-learning method — no model is fitted at training time; all computation happens at prediction time. This makes KNN impractical for real-time session scoring (as noted in the original paper), but it provides a useful non-parametric reference. We tune $k$ via cross-validation. Given the curse of dimensionality, we also try KNN on PCA-reduced features.

```{r knn-fit}
knn_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

cat("KNN tuning results:\n")
print(knn_model$results)

knn_pred <- predict(knn_model, test_x_scaled)
knn_prob <- predict(knn_model, test_x_scaled, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_y, positive = "Yes")
cat("\nKNN Test Results (best k =", knn_model$bestTune$k, "):\n")
print(knn_cm)

results[["KNN"]] <- list(
  model = knn_model,
  cm = knn_cm,
  probs = knn_prob
)
```

```{r knn-pca, fig.width=7, fig.height=5}
knn_pca_model <- train(
  x = train_pca_full,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pca_pred <- predict(knn_pca_model, test_pca_full)
knn_pca_prob <- predict(knn_pca_model, test_pca_full, type = "prob")

knn_pca_cm <- confusionMatrix(knn_pca_pred, test_y, positive = "Yes")
cat("KNN (7 PCs) Test Results (best k =", knn_pca_model$bestTune$k, "):\n")
print(knn_pca_cm)

results[["KNN (PCA)"]] <- list(
  model = knn_pca_model,
  cm = knn_pca_cm,
  probs = knn_pca_prob
)
```

## Model Comparison

We now compare all models on key metrics computed from their test set predictions.

```{r model-comparison, fig.width=9, fig.height=5}
library(pROC)

comparison <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  F1 = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (name in names(results)) {
  cm <- results[[name]]$cm
  probs <- results[[name]]$probs

  roc_obj <- roc(test_y, probs$Yes, levels = c("No", "Yes"), quiet = TRUE)

  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1 <- ifelse(is.na(prec) | (sens + prec) == 0, 0, 2 * sens * prec / (sens + prec))

  comparison <- rbind(comparison, data.frame(
    Model = name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(sens, 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1 = round(f1, 4),
    AUC = round(auc(roc_obj), 4),
    stringsAsFactors = FALSE
  ))
}

rownames(comparison) <- NULL
knitr::kable(comparison, caption = "Model Comparison on Test Set")
```

```{r roc-curves, fig.width=8, fig.height=6}
roc_list <- list()
for (name in names(results)) {
  roc_list[[name]] <- roc(test_y, results[[name]]$probs$Yes,
                          levels = c("No", "Yes"), quiet = TRUE)
}

plot(roc_list[[1]], col = 1, main = "ROC Curves: Model Comparison")
for (i in 2:length(roc_list)) {
  lines(roc_list[[i]], col = i)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=", 
                       sapply(roc_list, function(r) round(auc(r), 3)), ")"),
       col = 1:length(roc_list), lwd = 2, cex = 0.8)
```

The comparison table and ROC curves allow us to assess which methods are most effective for this dataset. Several patterns emerge:

1. **LDA achieves the highest AUC** among the methods tested so far, benefiting from the strong linear signal in features like `PageValues` and `ExitRates`. Its Gaussian class-conditional assumption is reasonable for the standardized numeric features.

2. **All models show low sensitivity.** This is a direct consequence of the class imbalance ($\sim 85\%$ non-purchase): with the default 0.5 probability threshold, models are biased toward predicting the majority class. High specificity paired with low sensitivity means these classifiers are conservative — they rarely predict a purchase unless the evidence is overwhelming.

3. **QDA on PCA-reduced features** provides a useful contrast to LDA. By allowing each class its own covariance structure, QDA can capture differences in variance between purchasers and non-purchasers, but it requires dimensionality reduction to avoid singular covariance estimates.

4. **Naive Bayes** performs reasonably despite its strong independence assumption. The conditional independence violation (e.g., `BounceRates`/`ExitRates` correlation) degrades its probability calibration but does not destroy the ranking of observations.

5. **KNN suffers from the curse of dimensionality.** Even after removing near-zero-variance columns, the feature space remains high-dimensional, making Euclidean distance less discriminative. KNN on PCA-reduced features may partially mitigate this.

These results motivate several next steps: (1) applying oversampling (e.g., SMOTE) or class weighting to improve sensitivity, (2) fitting regularized models (Lasso, Ridge) and tree-based ensembles (Random Forest, Gradient Boosting) that may better handle the feature space, and (3) tuning the classification threshold to balance sensitivity and specificity according to the business cost of missing a potential purchaser.




