---
title: "STAT432 Final Project"
date: "2026-04-30"
output:
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Team 13**

**Group Members:**

Andrew Di, amdi2
Praveen
Gautham

## Abstract {-}

[TODO: write a brief abstract of at most 200 words. Should cover (1) the prediction task and dataset, (2) the modeling approaches compared, (3) the headline result on the test set, and (4) the recommended approach.]

## Member Contributions {-}

Andrew Di: Data Preprocessing and Visualization, Logistic Regression
Praveen
Gautham

\newpage

# Introduction of the Task and Literature Review

"Online Shoppers Purchasing Intention Dataset" provided by UCI Machine Learning Repository [here](https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset).

The dataset contains $12{,}330$ observations of online shopping sessions, with $17$ input variables and one categorical output variable, `Revenue`, indicating whether the session ended in a purchase. The classes are imbalanced, with only about $15\%$ of sessions resulted in a purchase. There is no missing data.

We divide the $17$ predictors into four conceptual groups:

- **Page Interaction Features**: the number of pages of each type visited in the session and the total time spent on them: `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, and `ProductRelated_Duration`. These are derived from URL information and updated in real time as the user moves between pages.

- **Google Analytics Metrics** — engagement-quality metrics from Google Analytics: `BounceRates` (the percentage of visitors who enter the site from a page and leave without triggering any further requests), `ExitRates` (the percentage of pageviews on a page that were the last in the session), and `PageValues` (the average value of a page that a user visited before completing a transaction).

- **Session Context Features** — time- and seasonality-related variables: `SpecialDay`, `Month`, and `Weekend`. The `SpecialDay` value reflects how close the visit is to a holiday such as Valentine's Day or Mother's Day; for Valentine's Day, for instance, it takes a nonzero value between February 2 and February 12 and peaks at $1$ on February 8.

- **User Technical Profile** — session-level user attributes: `OperatingSystems`, `Browser`, `Region`, `TrafficType`, and `VisitorType` (whether the user is a new or returning visitor).

Our goal is to build classification models that can effectively determine whether or not an online shopper will make a purchase based on the feature variables above.

## Literature Review

[TODO: brief review of prior work, including the original paper introducing this dataset (Sakar et al., 2018, *Neural Computing and Applications*) and any related work on session-level e-commerce conversion prediction or imbalanced binary classification techniques relevant to the methods we use.]

# Details of the Dataset, Summary Statistics, and Data Visualization

## Data Preprocessing and Visualization

We'll start by loading in our data.

```{r load-data, message=FALSE}
library(tidyverse)

# load data
shoppers <- read_csv("online_shoppers_intention.csv")
glimpse(shoppers)

```

After loading in the data, we'll check for any missing data and print a summary of our dataset.

```{r clean, message=FALSE}

# missing data check
print(colSums(is.na(shoppers)))

# separate numeric and categorical features
numeric_features <- c(
  "Administrative", "Administrative_Duration",
  "Informational", "Informational_Duration",
  "ProductRelated", "ProductRelated_Duration",
  "BounceRates", "ExitRates", "PageValues", "SpecialDay"
)
categorical_features <- c(
  "Month", "OperatingSystems", "Browser",
  "Region", "TrafficType", "VisitorType", "Weekend"
)

# summary
summary(shoppers[, numeric_features])

```

We find that we don't have any missing data values. Additionally, we have the summary statistics for our numeric features above. Initially, we notice that most of our numeric features are concentrated near lower values but seem to have a right skew. We'll plot our features individually to visualize and identify any skewness or distributions we can take advantage of. We separate variables into numeric and categorical groups, plotting histograms for numeric features and bar plots for categorical features.

```{r univariate-numeric, fig.width=9, fig.height=7}
# Histogram of each numeric feature
par(mfrow = c(4, 3), mar = c(4, 4, 2, 1))
for (feat in numeric_features) {
  hist(shoppers[[feat]],
       breaks = 40,
       col = "steelblue",
       border = "white",
       main = feat,
       xlab = NULL)
}
par(mfrow = c(1, 1))  # reset plotting layout
```

After plotting, we confirm that each of our numeric features is mostly concentrated at low values and skewed to the right. We may want to perform some log transformation of our features to handle the skew. Notably, `SpecialDay` contains discrete values which may require special handling.

```{r univariate-categorical, fig.width=9, fig.height=7}
# Bar plot of level counts for each categorical feature
par(mfrow = c(3, 3), mar = c(6, 4, 2, 1))
for (feat in categorical_features) {
  counts <- table(as.character(shoppers[[feat]]))
  barplot(counts,
          col = "steelblue",
          main = feat,
          las = 2,         # vertical x-axis labels
          cex.names = 0.8)
}
par(mfrow = c(1, 1))
```

After plotting, we notice that several of our categorical features are heavily imbalanced. `Browser`, `OperatingSystems`, and `TrafficType` are mostly dominated by one or two levels. Similarly, `VisitorType` is mostly returning visitors with a smaller number of new visitors, and `Weekend` is mostly `FALSE`. `Month` shows a lot of seasonality with clear peaks in May and November, which could be explained by the end of the school year and graduation gifts in May, and the upcoming holiday season in November.

While plotting our features individually helps us understand their distribution, plotting the features against `Revenue` helps us identify if our feature distribution is dependent on whether or not a user made a purchase. If the conditional distributions of a feature given different `Revenue` value look different, then those features could carry more signal and are worth examining more. Below, we create box plots for our numeric features (compared against `Revenue`), and bar plots measuring per-level purchase rates for our categorical features. In our bar plots, the dashed line references the $\sim 15\%$ purchase rate in our overall dataset to compare against.

```{r bivariate-numeric, fig.width=9, fig.height=7}
# Boxplots of each numeric feature split by Revenue (FALSE / TRUE)
par(mfrow = c(4, 3), mar = c(3, 4, 2, 1))
for (feat in numeric_features) {
  boxplot(shoppers[[feat]] ~ shoppers$Revenue,
          col = c("salmon", "steelblue"),
          main = feat,
          xlab = "Revenue",
          ylab = NULL,
          outcex = 0.4)
}
par(mfrow = c(1, 1))
```

When visually inspecting the conditional distribution of our numeric features against `Revenue`, we find that when `Revenue` is `TRUE`, our `Administrative`, `Administrative_Duration`, `Informational`, `Informational_Duration`, `ProductRelated`, `ProductRelated_Duration`, and `PageValues` tend to be higher, while `BounceRates` and `ExitRates` tend to be lower. Because of the distinct distributions when compared against different values of `Revenue`, these might be features that are more important in our models.

Additionally, each of our box plots contains some amount of outliers. We believe the heavy right skewness of our features makes sense. For example, most users probably don't spend that much time on a page, which is why `Administrative_Duration` and `Informational_Duration` are mostly concentrated at low values. However, a few users will certainly spend far more time on a page. We decided to keep the outliers for each of our numeric features because we believe they capture and represent the skewed behavior that we've described.

```{r bivariate-categorical, fig.width=9, fig.height=7}
# Purchase rate within each level of each categorical feature.
# Dashed red line marks the overall ~15% purchase rate as a reference.
overall_rate <- mean(shoppers$Revenue)
par(mfrow = c(3, 3), mar = c(6, 4, 2, 1))
for (feat in categorical_features) {
  rates <- tapply(shoppers$Revenue, as.character(shoppers[[feat]]), mean)
  barplot(rates,
          col = "steelblue",
          main = feat,
          ylab = "Purchase Rate",
          las = 2,
          cex.names = 0.8,
          ylim = c(0, max(c(rates, overall_rate)) * 1.1))
  abline(h = overall_rate, lty = 2, col = "red")
}
par(mfrow = c(1, 1))
```

Now, when examining our categorical features, we can identify certain levels in our features that correspond with a larger purchase rate than average. For example, the purchase rate on the weekend is higher than average, and new visitors tend to have a higher purchase rate than average (and when compared to returning visitors). Users also tend to have a higher-than-average purchase rate in November, October, and September.

So far, we've examined each feature individually and compared against `Revenue`. It would also be useful for us to compare features against each other to identify any collinearity. Thus, we create a scatterplot matrix of our numeric features to identify pairwise relationships.

```{r scatterplot-matrix, fig.width=10, fig.height=10}
shoppers_numeric <- subset(shoppers, select = numeric_features)
pairs(shoppers_numeric,
      main = "Scatterplot Matrix for Numeric Features",
      col = "navy",
      pch = 20)
```

A few clear patterns stand out from the matrix. First, our corresponding count-and-duration feature pairs lie almost on a line: `Administrative` with `Administrative_Duration`, `Informational` with `Informational_Duration`, and `ProductRelated` with `ProductRelated_Duration`. This makes sense because visiting more pages of a given type leads to more total time spent on those types of pages. We also notice that `BounceRates` and `ExitRates` show a strong positive linear relationship, which makes sense given that both measure forms of leaving a page. Because of the collinearity between some of our features, we'll consider regularization in our models to handle redundancy.

# Classification Models

## Data Preparation for Modeling

Before fitting any models, we need to encode categorical variables, split the data into training and test sets, and standardize numeric features. We convert `Revenue` to a factor, encode `Month` and `VisitorType` as dummy variables, and treat the categorical features with numeric labels (`OperatingSystems`, `Browser`, `Region`, `TrafficType`) as factors to be dummy-encoded as well. Additionally, we set a seed to ensure reproducibility.

```{r data-prep, message=FALSE}
# libraries
library(caret)
library(MASS)
library(class)
library(e1071)
library(cluster)

# set seed
set.seed(432)

# convert Revenue to a factor with explicit "No"/"Yes" labels
# convert all categorical features to factors
shoppers_model <- shoppers
shoppers_model$Revenue <- factor(shoppers_model$Revenue,
                                 levels = c(FALSE, TRUE),
                                 labels = c("No", "Yes"))
factor_cols <- c("Month", "VisitorType", "Weekend",
                 "OperatingSystems", "Browser", "Region", "TrafficType")
for (col in factor_cols) {
  shoppers_model[[col]] <- factor(shoppers_model[[col]])
}

dummy_formula <- dummyVars(Revenue ~ ., data = shoppers_model, fullRank = TRUE)
shoppers_encoded <- as.data.frame(predict(dummy_formula, newdata = shoppers_model))
shoppers_encoded$Revenue <- shoppers_model$Revenue

train_idx <- createDataPartition(shoppers_encoded$Revenue, p = 0.7, list = FALSE)
train_data <- shoppers_encoded[train_idx, ]
test_data <- shoppers_encoded[-train_idx, ]

# Split predictors (X) from the response (y) for both train and test sets
train_x <- train_data[, names(train_data) != "Revenue"]
train_y <- train_data$Revenue
test_x <- test_data[, names(test_data) != "Revenue"]
test_y <- test_data$Revenue

nzv_cols <- nearZeroVar(train_x, names = TRUE)
if (length(nzv_cols) > 0) {
  cat("Removing near-zero variance columns:", paste(nzv_cols, collapse = ", "), "\n")
  train_x <- train_x[, !names(train_x) %in% nzv_cols]
  test_x <- test_x[, !names(test_x) %in% nzv_cols]
}

# and SpecialDay (discrete, only takes a handful of values).
# log-transform outputs to handle skew.
# skip BounceRates and ExitRates (bounded from 0 to 100)
# and SpecialDay (discrete)
log_features <- c("Administrative", "Administrative_Duration",
                  "Informational", "Informational_Duration",
                  "ProductRelated", "ProductRelated_Duration",
                  "PageValues")
train_x_log <- train_x
test_x_log  <- test_x
for (col in log_features) {
  if (col %in% names(train_x_log)) {
    train_x_log[[col]] <- log1p(train_x_log[[col]])
    test_x_log[[col]]  <- log1p(test_x_log[[col]])
  }
}

preproc <- preProcess(train_x, method = c("center", "scale"))
train_x_scaled <- predict(preproc, train_x)
test_x_scaled <- predict(preproc, test_x)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
cat("Training class distribution:\n")
print(table(train_y))
cat("Number of features after encoding:", ncol(train_x), "\n")
```

We split the dataset such that the training set consists of $70\%$ of the observations and the testing set consists of $30\%$ of the observations. We'll also log-transform the data so our models can handle the skewness.

## Logistic Regression

We fit four logistic regression variants: a default logistic regression without any penalty as baseline, Lasso ($L_1$), Ridge ($L_2$), and Elastic Net (a mix of both). To account for our skewness, we'll be using our log-transformed features rather than their un-transformed counterparts. The penalized variants are motivated by the multicollinearity we observed between count/duration pairs and between `BounceRates` and `ExitRates`. For each penalized variant we use `cv.glmnet` with 10-fold cross-validation to pick $\lambda$, and for Elastic Net we additionally grid-search over $\alpha$.

```{r logistic-setup, message=FALSE}
library(pROC)

# helper function that prints accuracy metrics for our
# logistic regression models
report_metrics <- function(name, prob, truth) {
  # Threshold probabilities at 0.5 to get predicted classes
  pred <- factor(ifelse(prob > 0.5, "Yes", "No"), levels = c("No", "Yes"))
  cm   <- confusionMatrix(pred, truth, positive = "Yes")

  # F1 from precision (Pos Pred Value) and recall (Sensitivity)
  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1   <- ifelse(is.na(prec) || (sens + prec) == 0, 0,
                 2 * sens * prec / (sens + prec))

  # AUC from the ROC curve
  auc_val <- as.numeric(auc(roc(truth, prob, levels = c("No", "Yes"),
                                 quiet = TRUE)))

  cat(sprintf("%s:\n", name))
  cat(sprintf("  Accuracy:    %.4f\n", cm$overall["Accuracy"]))
  cat(sprintf("  Sensitivity: %.4f\n", sens))
  cat(sprintf("  Specificity: %.4f\n", cm$byClass["Specificity"]))
  cat(sprintf("  F1:          %.4f\n", f1))
  cat(sprintf("  AUC:         %.4f\n", auc_val))
}
```

### Baseline (Unpenalized) Logistic Regression

We start with a standard logistic regression on all encoded predictors as a benchmark.

```{r logistic-baseline, message=FALSE, warning=FALSE}
set.seed(432)

# training and testing dataframes (log-transformed numeric features)
log_train_df <- data.frame(Revenue = train_y, train_x_log)
log_test_df  <- data.frame(Revenue = test_y,  test_x_log)

# Fit default logistic regression on all features
log_baseline <- glm(Revenue ~ ., data = log_train_df, family = binomial())

# Predicted P(Revenue = "Yes") on the test set
log_baseline_prob <- predict(log_baseline, newdata = log_test_df, type = "response")

# Print headline test-set metrics
report_metrics("Baseline Logistic Regression", log_baseline_prob, test_y)
```

Using our baseline Logistic Regression model, we find the metrics above. Notably, our Accuracy of $0.8902$ is higher than a naive accuracy of around $0.85$. However, our Sensitivity is really low at $0.5542$.

### Lasso-Penalized Logistic Regression ($L_1$)
Adding a Lasso penalty shrinks coefficients toward zero, which helps us perform feature selection. This helps us select the most important features while reducing model complexity and noise. We use `cv.glmnet` with 10-fold CV to pick the $\lambda$ that minimizes cross-validated deviance.

```{r logistic-lasso, message=FALSE}
set.seed(432)
library(glmnet)

# matrix setup (log-transformed numeric features)
x_train_mat <- as.matrix(train_x_log)
x_test_mat  <- as.matrix(test_x_log)

# Cross-validation over lambda for Lasso (alpha = 1)
lasso_cv <- cv.glmnet(x = x_train_mat, y = train_y,
                      family = "binomial", alpha = 1, nfolds = 10)
cat("Lasso lambda.min:", lasso_cv$lambda.min, "\n")

# use lambda.min to make predictions
log_lasso_prob <- as.numeric(predict(lasso_cv, newx = x_test_mat,
                                     s = "lambda.min", type = "response"))

report_metrics("Lasso Logistic Regression", log_lasso_prob, test_y)
```
After performing our cross-validation, we use $\lambda = 0.001526258$. By using our Lasso penalty, we get very small improvements in every metric over our baseline logistic regression model.

### Ridge-Penalized Logistic Regression ($L_2$)

Adding a Ridge penalty shrinks all coefficients toward zero but never to exactly zero, which helps handle the collinearity we have between our features. We use the same cross-validation procedure as Lasso, but with `alpha = 0`.

```{r logistic-ridge, message=FALSE}
set.seed(432)

# Cross-validation over lambda for Ridge (alpha = 0)
ridge_cv <- cv.glmnet(x = x_train_mat, y = train_y,
                      family = "binomial", alpha = 0, nfolds = 10)
cat("Ridge lambda.min:", ridge_cv$lambda.min, "\n")

log_ridge_prob <- as.numeric(predict(ridge_cv, newx = x_test_mat,
                                     s = "lambda.min", type = "response"))

report_metrics("Ridge Logistic Regression", log_ridge_prob, test_y)
```

After performing our cross-validation, we use $\lambda = 0.02319776$. However, when we use our Ridge penalty, our model performs worse than our  Lasso-penalized Logistic Regression model in every metric except for Specificity and AUC, in which it performs slightly better.

### Elastic Net Logistic Regression

Elastic Net blends $L_1$ and $L_2$ via a mixing parameter $\alpha \in (0, 1)$. This gives us the benefit of being able to use feature selection and handle multicollinearity, at the cost of computational complexity due to the additional hyperparameter that we need to tune.
`cv.glmnet` only tunes $\lambda$, so we manually grid-search over $\alpha$ and pick the combination with the lowest cross-validated error.

```{r logistic-elastic, message=FALSE}
set.seed(432)

# search for alpha between 0.1 to 0.9, stepsize 0.1
alpha_grid <- seq(0.1, 0.9, by = 0.1)
elastic_cv_results <- lapply(alpha_grid, function(a) {
  set.seed(432)  # same fold structure across alphas for a fair comparison
  cv.glmnet(x = x_train_mat, y = train_y,
            family = "binomial", alpha = a, nfolds = 10)
})

# Pick the alpha whose best lambda gives the lowest CV error
cv_min_errors  <- sapply(elastic_cv_results, function(fit) min(fit$cvm))
best_alpha_idx <- which.min(cv_min_errors)
best_alpha     <- alpha_grid[best_alpha_idx]
elastic_cv     <- elastic_cv_results[[best_alpha_idx]]

cat("Elastic Net best alpha:", best_alpha,
    "; lambda.min:", elastic_cv$lambda.min, "\n")

log_elastic_prob <- as.numeric(predict(elastic_cv, newx = x_test_mat,
                                       s = "lambda.min", type = "response"))

report_metrics("Elastic Net Logistic Regression", log_elastic_prob, test_y)
```

After performing our grid search and cross-validation, we find that our best hyperparameters are $\alpha = 0.9$ and $\lambda = 0.001545188$. Using these hyperparameters, our Elastic Net model performs similarly to our Lasso-penalized model, performing slightly worse in Sensitivity, F1, and AUC, slightly better in Specificity, and the same in Accuracy.

As a whole, our logistic regression model performs similarly well when using a Lasso penalty and in Elastic Net. However, Elastic Net required a lot more tuning for negligibly better or worse results, so our Lasso-penalized model would likely be the best Logistic Regression model to use moving forward.

## Unsupervised Learning

### Principal Component Analysis (PCA)

We apply PCA to the 10 original numeric features (before dummy encoding) to understand the variance structure and to provide reduced-dimension inputs for later models.

```{r pca, fig.width=9, fig.height=5}
train_numeric <- train_data[, numeric_features]
test_numeric <- test_data[, numeric_features]

pca_preproc <- preProcess(train_numeric, method = c("center", "scale"))
train_numeric_scaled <- predict(pca_preproc, train_numeric)
test_numeric_scaled <- predict(pca_preproc, test_numeric)

pca_fit <- prcomp(train_numeric_scaled, center = FALSE, scale. = FALSE)

var_explained <- summary(pca_fit)$importance
par(mfrow = c(1, 2))
plot(var_explained[2, ], type = "b", pch = 19, col = "steelblue",
     xlab = "Principal Component", ylab = "Proportion of Variance",
     main = "Scree Plot")
plot(var_explained[3, ], type = "b", pch = 19, col = "darkred",
     xlab = "Principal Component", ylab = "Cumulative Proportion",
     main = "Cumulative Variance Explained")
abline(h = 0.9, lty = 2, col = "gray40")
par(mfrow = c(1, 1))

cat("Variance explained by first 5 PCs:", round(var_explained[3, 5], 3), "\n")
cat("Variance explained by first 7 PCs:", round(var_explained[3, 7], 3), "\n")
```

```{r pca-biplot, fig.width=8, fig.height=6}
train_pca_scores <- as.data.frame(pca_fit$x[, 1:2])
train_pca_scores$Revenue <- train_y

ggplot(train_pca_scores, aes(x = PC1, y = PC2, color = Revenue)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "PCA: First Two Principal Components Colored by Revenue",
       x = paste0("PC1 (", round(var_explained[2, 1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(var_explained[2, 2] * 100, 1), "%)")) +
  theme_minimal() +
  scale_color_manual(values = c("No" = "gray60", "Yes" = "tomato"))
```

The PCA biplot shows that the two classes overlap substantially in the first two principal components, indicating that separating purchasers from non-purchasers requires more than two linear directions. However, `Revenue = Yes` sessions tend to have a wider spread, particularly along PC1, suggesting that high-engagement sessions (which load onto the page count and duration variables) are more likely to result in purchases.

### K-Means Clustering

We cluster the scaled numeric features to discover natural session groupings. We use the silhouette method to choose the number of clusters.

```{r kmeans-silhouette, fig.width=7, fig.height=5}
set.seed(432)
sample_idx <- sample(nrow(train_numeric_scaled), 3000)
train_sample <- train_numeric_scaled[sample_idx, ]

sil_widths <- sapply(2:8, function(k) {
  km <- kmeans(train_sample, centers = k, nstart = 25, iter.max = 50)
  ss <- silhouette(km$cluster, dist(train_sample))
  mean(ss[, 3])
})

plot(2:8, sil_widths, type = "b", pch = 19, col = "steelblue",
     xlab = "Number of Clusters (k)", ylab = "Average Silhouette Width",
     main = "Silhouette Analysis for K-Means")
best_k <- which.max(sil_widths) + 1
cat("Best k by silhouette:", best_k, "\n")
```

```{r kmeans-fit, fig.width=9, fig.height=5}
km_fit <- kmeans(train_numeric_scaled, centers = best_k, nstart = 25, iter.max = 100)

train_pca_all <- as.data.frame(pca_fit$x[, 1:2])
train_pca_all$Cluster <- factor(km_fit$cluster)
train_pca_all$Revenue <- train_y

p1 <- ggplot(train_pca_all, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "K-Means Clusters in PCA Space") +
  theme_minimal()

# Per-cluster sample size and purchase rate (for the bar plot below)
cluster_n <- table(train_pca_all$Cluster)
cluster_rates <- tapply(train_pca_all$Revenue == "Yes",
                        train_pca_all$Cluster, mean)
cluster_revenue <- data.frame(
  Cluster = factor(names(cluster_n), levels = names(cluster_n)),
  n = as.numeric(cluster_n),
  purchase_rate = as.numeric(cluster_rates)
)

p2 <- ggplot(cluster_revenue, aes(x = Cluster, y = purchase_rate, fill = Cluster)) +
  geom_col() +
  geom_hline(yintercept = mean(train_y == "Yes"), linetype = "dashed") +
  labs(title = "Purchase Rate by Cluster",
       subtitle = "Dashed line = overall training rate",
       y = "Purchase Rate") +
  theme_minimal() +
  theme(legend.position = "none")

gridExtra::grid.arrange(p1, p2, ncol = 2)
print(cluster_revenue)
```

```{r kmeans-profiles}
# Mean of each numeric feature within each cluster (original scale, not scaled).
# aggregate() returns one row per cluster with feature-mean columns.
cluster_profiles <- aggregate(
  train_data[, numeric_features],
  by = list(Cluster = factor(km_fit$cluster)),
  FUN = mean
)

# Transpose so rows are features and columns are clusters (easier to scan in a table).
cluster_means_t <- t(cluster_profiles[, -1])
colnames(cluster_means_t) <- paste("Cluster", cluster_profiles$Cluster)
cluster_profiles_long <- data.frame(
  Feature = rownames(cluster_means_t),
  cluster_means_t,
  check.names = FALSE,
  row.names = NULL
)

knitr::kable(cluster_profiles_long, digits = 2, caption = "Cluster Centroids (Original Scale)")
```

### Hierarchical Clustering

We also apply hierarchical clustering (Ward's method) to a subsample and display a dendrogram to reason about the natural number of groups.

```{r hierarchical, fig.width=9, fig.height=5}
hc_dist <- dist(train_sample)
hc_fit <- hclust(hc_dist, method = "ward.D2")

plot(hc_fit, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's Method, n=3000 sample)",
     xlab = "", sub = "")
rect.hclust(hc_fit, k = best_k, border = "steelblue")
```

The dendrogram corroborates the K-means result: the data naturally splits into a small number of groups, broadly distinguishing low-engagement sessions from high-engagement ones.

## Supervised Learning

We now turn to classification. For all supervised models, we evaluate using 5-fold stratified cross-validation on the training set. Because the classes are imbalanced ($\sim 85\%$ non-purchase), we report not just accuracy but also sensitivity (recall for purchases), specificity, and AUC. We collect results into a summary table at the end.

```{r cv-setup}
# trainControl is used by the caret-based models below (LDA, KNN).
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

results <- list()
```

### Discriminant Analysis: Choosing Between LDA, QDA, and Naive Bayes

Among generative classifiers, we consider three options: Linear Discriminant Analysis (LDA), Quadratic Discriminant Analysis (QDA), and Naive Bayes. We select **LDA** for the following reasons:

- **QDA** requires estimating a separate covariance matrix for each class. With our high-dimensional dummy-encoded feature space, the within-class covariance matrices become singular or near-singular, forcing us to restrict QDA to a small PCA subspace and lose information from the categorical predictors.
- **Naive Bayes** assumes conditional independence among features given the class — a strong assumption clearly violated here (e.g., `BounceRates` and `ExitRates` are highly correlated, as are page counts and durations). While Naive Bayes can still produce reasonable rankings, its probability estimates are poorly calibrated under dependence, and it typically underperforms LDA when the shared-covariance assumption is approximately satisfied.
- **LDA** assumes multivariate normality with a shared covariance matrix across classes. While normality is imperfect for our skewed features, the shared-covariance assumption is more defensible than independence, and LDA can operate on the full encoded feature set without dimensionality issues. It also provides a natural projection onto the discriminant axis, aiding interpretation.

### Linear Discriminant Analysis (LDA)

LDA projects the data onto the single linear discriminant axis that maximally separates the two classes. Given the collinearity we observed earlier, we should expect LDA to perform comparably to logistic regression.

```{r lda-fit}
lda_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "lda",
  trControl = ctrl,
  metric = "ROC"
)

lda_pred <- predict(lda_model, test_x_scaled)
lda_prob <- predict(lda_model, test_x_scaled, type = "prob")

lda_cm <- confusionMatrix(lda_pred, test_y, positive = "Yes")
cat("LDA Test Results:\n")
print(lda_cm)

results[["LDA"]] <- list(
  model = lda_model,
  cm = lda_cm,
  probs = lda_prob
)
```

### K-Nearest Neighbors (KNN)

KNN classifies each observation by a majority vote of its $k$ nearest neighbors. It is a non-parametric, lazy-learning method — no model is fitted at training time; all computation happens at prediction time. This makes KNN impractical for real-time session scoring (as noted in the original paper), but it provides a useful non-parametric reference. We tune $k$ via cross-validation. Given the curse of dimensionality, we also try KNN on PCA-reduced features.

```{r knn-fit}
knn_model <- train(
  x = train_x_scaled,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

cat("KNN tuning results:\n")
print(knn_model$results)

knn_pred <- predict(knn_model, test_x_scaled)
knn_prob <- predict(knn_model, test_x_scaled, type = "prob")

knn_cm <- confusionMatrix(knn_pred, test_y, positive = "Yes")
cat("\nKNN Test Results (best k =", knn_model$bestTune$k, "):\n")
print(knn_cm)

results[["KNN"]] <- list(
  model = knn_model,
  cm = knn_cm,
  probs = knn_prob
)
```

```{r knn-pca, fig.width=7, fig.height=5}
train_pca_full <- as.data.frame(pca_fit$x[, 1:7])
test_pca_full <- as.data.frame(predict(pca_fit, test_numeric_scaled)[, 1:7])

knn_pca_model <- train(
  x = train_pca_full,
  y = train_y,
  method = "knn",
  trControl = ctrl,
  metric = "ROC",
  tuneGrid = data.frame(k = c(3, 5, 7, 11, 15, 21, 31, 51))
)

knn_pca_pred <- predict(knn_pca_model, test_pca_full)
knn_pca_prob <- predict(knn_pca_model, test_pca_full, type = "prob")

knn_pca_cm <- confusionMatrix(knn_pca_pred, test_y, positive = "Yes")
cat("KNN (7 PCs) Test Results (best k =", knn_pca_model$bestTune$k, "):\n")
print(knn_pca_cm)

results[["KNN (PCA)"]] <- list(
  model = knn_pca_model,
  cm = knn_pca_cm,
  probs = knn_pca_prob
)
```

## Model Comparison

We now compare all models on key metrics computed from their test set predictions.

```{r model-comparison, fig.width=9, fig.height=5}
library(pROC)

comparison <- data.frame(
  Model = character(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  F1 = numeric(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

for (name in names(results)) {
  cm <- results[[name]]$cm
  probs <- results[[name]]$probs

  roc_obj <- roc(test_y, probs$Yes, levels = c("No", "Yes"), quiet = TRUE)

  sens <- cm$byClass["Sensitivity"]
  prec <- cm$byClass["Pos Pred Value"]
  f1 <- ifelse(is.na(prec) | (sens + prec) == 0, 0, 2 * sens * prec / (sens + prec))

  comparison <- rbind(comparison, data.frame(
    Model = name,
    Accuracy = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(sens, 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1 = round(f1, 4),
    AUC = round(auc(roc_obj), 4),
    stringsAsFactors = FALSE
  ))
}

rownames(comparison) <- NULL
knitr::kable(comparison, caption = "Model Comparison on Test Set")
```

```{r roc-curves, fig.width=8, fig.height=6}
roc_list <- list()
for (name in names(results)) {
  roc_list[[name]] <- roc(test_y, results[[name]]$probs$Yes,
                          levels = c("No", "Yes"), quiet = TRUE)
}

plot(roc_list[[1]], col = 1, main = "ROC Curves: Model Comparison")
for (i in 2:length(roc_list)) {
  lines(roc_list[[i]], col = i)
}
legend("bottomright",
       legend = paste0(names(roc_list), " (AUC=",
                       sapply(roc_list, function(r) round(auc(r), 3)), ")"),
       col = 1:length(roc_list), lwd = 2, cex = 0.8)
```

# Conclusion

[TODO: summarize the main findings — which model performed best on the test set and why, what the headline AUC / sensitivity / specificity numbers were, what the most important predictors were, and what the recommended modeling approach would be for a production setting. Note any limitations (class imbalance handling, models not yet tried) and possible directions for future work.]
